### 2023-04-11 killer_plot1.Rmd
Add killer plot

---
title: "Killer Plot"
subtitle: "Group 3"
author: "Grace Wang, Virginia Baskin, Siddhi Narayan"
date: "2023-04-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSQLite)
library(dplyr)
library(stringr)
library(grid)
library(readr)
options(dplyr.summarise.inform = FALSE)

full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
mini_htx <- full_htx[seq(1,2000000,3),]
mini_htx <- mini_htx %>% rename("Beats" = "beat")
#beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
#htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")
```

```{r, echo=FALSE, results='hide'}

# subset the data to only contain columns of interest
citations_htx <- mini_htx[,c("subject_race", "subject_sex", "date", "violation")]

# subset the data to only contain rows of violation interest
citations_htx <- citations_htx %>% 
  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)|
           grepl("SEAT BELT", violation)|grepl("FINANCIAL RESPONSIBILITY", violation)|
           grepl("LICENSE", violation)|grepl("SPEEDING", violation)) %>%
  mutate(date = format(date, format = "%Y"))

#create the circles dataframe which counts individuals in each race, year pairing
citations_htx_circle <- citations_htx %>%
  group_by(subject_race, date) %>%
  summarise(n_individuals = n())
citations_htx_circle

# create the bars dataframe which counts the number of individuals in each race, year, violation, pairing

# create a dataframe with all possible combinations of race, date, sex, and violation
citations_htx_bar <- citations_htx %>%
  mutate(red_stop= if_else(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation), 1, 0)) %>%
  mutate(seat_belt= if_else(grepl("SEAT BELT", violation), 1, 0)) %>%
  mutate(fin_resp= if_else(grepl("FINANCIAL RESPONSIBILITY", violation), 1, 0))%>%
  mutate(license= if_else(grepl("LICENSE", violation), 1, 0))%>%
  mutate(speeding= if_else(grepl("SPEEDING", violation), 1, 0))%>%
  select(-violation)%>%
  group_by(date, subject_race, subject_sex, red_stop, seat_belt, fin_resp, license, speeding)%>%
  summarise(n_individuals = n())
citations_htx_bar
```

```{r Barplot Function, echo=FALSE, results='hide'}
barplot <- function(red_stop_m, red_stop_f, seat_belt_m, seat_belt_f, fin_resp_m, fin_resp_f, license_m, license_f, speeding_m, speeding_f){
  y <- c(0.5, 0.5)
  x <- c(0, 1.1)
  #grid.newpage()
  citations <- c("red_stop", "seat_belt", "fin_resp", "license", "speeding")
  male <- c(red_stop_m, seat_belt_m, fin_resp_m, license_m, speeding_m)
  max_male <- max(male)
  female <- c(red_stop_f, seat_belt_f, fin_resp_f, license_f, speeding_f)
  max_female <- max(female)
  colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
  for (i in 1:5){
    male_scale <-  male[i]/(2*max(male))
    female_scale <- female[i]/(2*max(female))
    grid.rect(x=0.2*i-0.1, y = 0.5+0.5*male_scale, height = male_scale, width = 0.199, gp = gpar(fill = colors[i]))
    grid.rect(x=0.2*i-0.1, y = 0.5-0.5*female_scale, height = female_scale, width = 0.199, gp = gpar(fill = colors[i]))
    #grid.text(citations[i], x = 0.2*i-0.1, y = 0.52)
  }
   grid.lines(x, y, gp = gpar(lwd = 1, col = "black"))
   #grid.text ("M", x = 0.9, y = 0.9)
   #grid.text ("F", x = 0.9, y = 0.1)
}
```

```{r slay, echo=FALSE, results='hide'}
bar_run <- function(plot_race,plot_year){
  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year), ]
  new_df <- new_df[which(citations_htx_bar$subject_race == plot_race),]
  male_df <- new_df[which(new_df$subject_sex == "male"),]
  female_df <- new_df[which(new_df$subject_sex == "female"),]
  red_stop_m <- sum(male_df$red_stop == 1)
  red_stop_f <- sum(female_df$red_stop == 1)
  seat_belt_m <- sum(male_df$seat_belt == 1)
  seat_belt_f <- sum(female_df$seat_belt == 1)
  fin_resp_m <- sum(male_df$fin_resp == 1)
  fin_resp_f <- sum(female_df$fin_resp == 1)
  license_m <- sum(male_df$license == 1)
  license_f <- sum(female_df$license == 1)
  speeding_m <- sum(male_df$speeding == 1)
  speeding_f <- sum(female_df$speeding == 1)
  barplot(red_stop_m, red_stop_f, seat_belt_m, seat_belt_f, fin_resp_m, fin_resp_f, license_m, license_f, speeding_m, speeding_f)
}
#bar_run("white", 2014)  
```

```{r Circles, echo=FALSE, results='hide'}

#DRAW CIRCLE FUNCTION DONE

#given frequency, plot a properly scaled circle in the center of the viewpoint
#max_cite is the citation count of the race, year with the most citations
#found by max(cite_counts_list)
draw_circle <- function(count, max_count = max_cite){
  my_circle <- circleGrob(name = "my_circle",
                          x = 0.5, y = 0.5, r = .8*count/(max_count), #radius scale needs to be fixed and them im done
                          gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.draw(my_circle)
}

# draw_circle(6, 10)
# popViewport()

```

```{r, echo=FALSE, results='hide'}
legend <- function(citations, colors){
  grid.rect(x = 0, y= 0, just = c("left", "bottom"), width = 1, height = .9)
  grid.text("LEGEND",
          x = .5, y = 0.915, just = c("center", "bottom"),
          gp = gpar(fontsize = 10, fontface = "bold"))

  #citations + corresponding colors
  #citations <- c("Red/Stop", "Seat Belt", "Financial", "License", "Speeding")
  #colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
  ys <- c(0.74, 0.65, 0.56, 0.47, 0.38)
  grid.text("Citation Type", x= 0.05, y= 0.87, just = c("left", "top"), gp = gpar(fontsize=8, fontface="bold"))
  for (i in 1:5){
    grid.rect(x = 0.17, y = ys[i], just = c("left", "bottom"), height = 0.077, width = 0.15, 
              gp = gpar(col = "black", fill=colors[i]))
    grid.text(citations[i], x = 0.37, y = ys[i]+0.038, just = c("left", "center"), gp = gpar(fontsize=7))
  }

  grid.text("Citation Frequency", x = 0.04, y = 0.34, just = c("left", "top"), gp = gpar(fontsize=8, fontface="bold"))
  grid.text("Compared to other (race, year)s", x = 0.037, y = 0.3, just = c("left", "top"), gp = gpar(fontsize=5))
  #comparatively less citations for race that year 
  #comparatively more citations

  grid.circle(x = 0.15, y = 0.233, r = 0.04, gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.text("Less Citations", x = 0.3, y = 0.233, just = c("left", "center"), gp = gpar(fontsize=7))
  grid.text("Overall", x = 0.43, y = 0.2, just = c("left", "center"), gp = gpar(fontsize=7))

  grid.circle(x = 0.18, y = 0.11, r = 0.175, gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.text("More Citations", x = 0.4, y = 0.11, just = c("left", "center"), gp = gpar(fontsize=7))
  grid.text("Overall", x = 0.53, y = 0.075, just = c("left", "center"), gp = gpar(fontsize=7))
}
```

```{r Putting Together the Killer Plot, echo=FALSE}
# PUT IT ALL TOGETHER
grid.newpage()

#legend viewport 
pushViewport(viewport(
  x = 0.84, y = 0.33, just = c("left", "bottom"),
  width = 0.16, height = 0.57
))

#plot legend
citations <- c("Red/Stop", "Seat Belt", "Financial", "License", "Speeding")
colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
legend(citations, colors)

popViewport()

#main viewport
pushViewport(viewport(
  x = 0.1, y = 0.1, just = c("left", "bottom"),
  width = 0.8, height = 0.9
))
#axis
grid.lines(c(1.75/8, 0.9), c(1.5/8, 1.5/8))
grid.lines(c(1.75/8, 1.75/8), c(1.5/8, 0.9))
#grid.lines(c(0.1, 0.9), c(0.5, 0.5))
grid.text("Killer Plot",
          x = 1.75/8, y = 0.925, just = c("left", "bottom"),
          gp = gpar(fontsize = 18)
)
pushViewport(viewport(
  x = 0.1, y = 0.1, just = c("left", "bottom"),
  width = 0.8, height = 0.8
))
#grid.rect(x=0.5, y=0.5, width = 1, height=1)
rows <- 8
cols <- 6
ly <- grid.layout(rows, cols)
pushViewport(viewport(layout = ly))
#makes a viewpoint for each

# vectors to iterate through for different parameters
races <- unique(citations_htx_bar$subject_race) # races
years <- unique(citations_htx_bar$date) # dates

# set row labels
for (i in 1:(rows-1)) {
  pushViewport(viewport(layout.pos.row=i, layout.pos.col = 1))
  grid.text(years[i])
  popViewport()}

# set column labels
for (j in 2:cols){
  pushViewport(viewport(layout.pos.row=8, layout.pos.col = j))
  if (races[j-1]=="asian/pacific islander"){
    grid.text("AAPI", gp = gpar(fontsize = 12))
  }
  else{
    grid.text(races[j-1], gp = gpar(fontsize = 12))
  }
  popViewport()
}

for (i in 1:(rows-1)) {
  for (j in 2:cols) {

    # push main grid Viewport
    pushViewport(viewport(layout.pos.row=i, layout.pos.col = j))

    # push small viewport
    pushViewport(viewport(width=0.75, height=0.75, just="centre"))

    # make bar plot
    bar_run(races[j-1], years[i])

    if (j==2){
      if (i==1){
        grid.text("M", x = 0.1, y = 0.9, gp=gpar(fontsize=8))
        grid.text("F", x = 0.1, y = 0.1, gp=gpar(fontsize=8))
      }
    }

    # make circle
    count <- pull(citations_htx_circle[citations_htx_circle['subject_race']==races[j-1] &                          citations_htx_circle['date']==years[i],"n_individuals"])[1]
    #max_count <- sum(citations_htx_circle$n_individuals)
    max_count <- max(citations_htx_circle$n_individuals)
    draw_circle(count, max_count)

    # pop small viewport
    popViewport()

    # pop main grid viewport
    popViewport()
  }
}
```

For our killer plot, we have constructed a visualization depicting the frequency of each citation type for every race, year combination. The frequency of citations is represented by mirrored bar plots for each year and race combination, with citation count for males being plotted above the horizontal axis and citation count for females being plotted below. Overlaid on each bar plot is a circle, the radius of which is normalized to the race, year combination with the greatest citations, therefore providing a comparative representation of the number of citations for each race, year combination. Altogether, this plot allows us to comprehensively view our dataset and the distribution of citations across race, year, and sex.

From this plot, we observed a fairly symmetrical distribution of citations between males in females in all years for white and black individuals. However, for individuals of AAPI, individual, and unknown race, the distribution of arrests fluctuated between each gender each year, possibly due to the low number of arrests for these three races overall. Notably, in 2015, 2018, and 2020, it appears that nearly all citations issued to AAPI individuals were given to women, but the low overall quantity of citations given to AAPI individuals in these years means that this pattern is likely not indicative of any bias.

Among black, white and AAPI individuals, license-related issues and failure to establish financial responsibility were the two most common citations in every year. This makes sense, as these two citations are often concurrent with others, as officers must check individuals' license and registration after being pulled over for any offense.

### 2023-04-11 draft4.Rmd
Add latest draft of report

---
title: "Investigating Police Citations in Houston"
subtitle: "Draft 4, Group 3"
author: "Grace Wang, Virginia Baskin, Siddhi Narayan"
date: "2023-03-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load packages, include=FALSE}
library(RSQLite)
library(readr)
library(ggplot2)
library(ggExtra)
library(ggthemes)
library(dplyr)

mini_htx <- read_csv("Datasets/mini_htx.csv")
full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
mini_htx <- full_htx[seq(1,2000000,3),]
mini_htx <- mini_htx %>% rename("Beats" = "beat")
beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")
#dbWriteTable(con, "census", census_csv, overwrite = TRUE)

```

## Introduction

Our team's dataset is sourced from the Stanford Open Policing Project and contains information on police stops that result in citations in Houston, TX from 2014-2020. The data contains a little over 2 million rows. The fields of the dataset include (non-exhaustively): the date, time, location, latitude, longitude of the stop, the police beat of the officer, district, the subject's race, sex, the type of violation they received, and the vehicle make, model, and color. Our team approached this dataset with a special interest in investigating trends regarding subject race, sex, and speeding amounts. Ultimately, we seek to investigate any factors that particularly dispose individuals towards receiving a citation.

## Data Exploration - Graphs

We utilized various visualization techniques in order to comprehensively analyze every feature of the problem space and maximize various perspectives on the dataset. Our dataset has a lot of free text, which is difficult to visualize in the typical way. We visualized every field that was not unstructured free text in some way or another.

```{r Race Line Graph by Year, echo=FALSE}
# Visualize totals for each race by year
# Grace

race_date <- mini_htx[,c("date", "subject_race")] # race and date
race_date <- na.omit(race_date) # omit na Values

year_only <- function(fulldate){
  year <- substring(fulldate, 1, 4) # only keep year
}

black <- race_date[race_date$subject_race=="black",]
black$year <- unlist(lapply(black$date, year_only))
plot(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), ylim = c(0, 60000), main = "Citations by Race for Each Year", xlab="Year", ylab = "Count") #200000 when all
lines(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), type = "l", col = "blue")

white <- race_date[race_date$subject_race=="white",]
white$year <- unlist(lapply(white$date, year_only))
points(sort(unique(white$year)), as.vector(table(as.factor(white$year))))
lines(sort(unique(white$year)), as.vector(table((as.factor(white$year)))), type = "l", col = "red")

aapi <- race_date[race_date$subject_race=="asian/pacific islander",]
aapi$year <- unlist(lapply(aapi$date, year_only))
points(sort(unique(aapi$year)), as.vector(table(as.factor(aapi$year))))
lines(sort(unique(aapi$year)), as.vector(table((as.factor(aapi$year)))), type = "l", col = "green")

unknown <- race_date[race_date$subject_race=="unknown",]
unknown$year <- unlist(lapply(unknown$date, year_only))
points(sort(unique(unknown$year)), as.vector(table(as.factor(unknown$year))))
lines(sort(unique(unknown$year)), as.vector(table((as.factor(unknown$year)))), type = "l", col = "orange")

legend("topright", legend=c("black", "white", "aapi", "unknown"), col=c("blue", "red", "green", "orange"), pch = 20, cex = 0.7, title = "Race")

```

First, we turn our attention to the total individuals of each race issued citations in each year 2014-2020. From 2014-2019, the greatest number of individuals issued citations were white, but in 2020, black individuals narrowly surpassed white individuals as the racial group with the most citations of any racial group. Overall, individuals of unknown race and asian/pacific islanders received significantly fewer citations than white and black individuals. In 2019 and 2020, a decrease in citations issued to white individuals can be observed. This data is not sufficient for drawing conclusions of racial bias in the citation process; more investigation is necessary into confounding factors.

```{r Race to citation ratio, echo=FALSE}
df <- data.frame(table(na.omit(mini_htx$subject_race)))
total_freq <- sum(df$Freq)
df$prop <- df$Freq/total_freq
num_squares <- round(df$prop*100)

df_grid <- data.frame(
 x = rep(1:10, each = 10),
 y = rep(1:10, times = 10)
)
num_squares <- c(3, 36, 1, 3, 57) #manually add one more square because the rounding resulted in only 99 
#num_squares <- round(df$prop*100)

df_grid$category <- rep(df$Var1, num_squares)
ggplot(df_grid, aes(x = x, y = y, fill = category)) +
 geom_tile(color = "black", size = 0.5) +
 scale_fill_brewer(palette = "Set3") +
 labs(title="Waffle Chart of Citations per Race")+
 theme_void()
```

Next, we will look at the % of Citation by Race across all years 2014-2020 in a more geometric representation of the percentages. As we can see, white individuals have the highest percentage of citations, followed by black individuals, and then Asian/pacific islanders, then unknown races. This is somewhat proportional to the actual demographics of Houston (according to the U.S. census) – the White and Black percentages of citations are slightly higher than their population in Houston, while the proportion of citations for Asian individuals is slightly lower than the proportion of Asian people in Houston. This graph is a fun (and new, for us) way to visualize which races make up what proportions in our dataset.

```{r Line graph of number of citations, echo=FALSE}
# Observe citations over time and any time-associate patterns
time <- as.POSIXct(mini_htx$time, format = "%H:%M:%S")
hour <- as.numeric(format(time, "%H"))
count <- data.frame(table(na.omit(hour)))
plot(count$Var1, count$Freq, type = "n", xlab = "Hour", ylab = "Frequency of Citations", main = "Frequency of Citations vs Hour in Day")
lines(count$Var1, count$Freq, type = "l", lwd = 2)
```

Now we will look at the Frequency of citations over the Hours of the Day. Citations are most common around 8 am, and around 3 pm. There may be a higher proportion of citations around 8 am because of high travel due to work. We are not quite sure why another peak is found at 3 pm, it may be people returning from school and/or work. Children in K-12 grade usually get out of school around 3pm. This peak could be correlated with people speeding to pick up/return home with their children before rush hour. We also see less citations during the super late/super early hours –  likely due to less travel as a whole.

```{r Scatter plot longitude vs latitude by district, echo=FALSE}
# Verifying the hubs for citation and locations of districts
mini_htx2 <- mini_htx[seq(1,666666,30),]
mini_htx2$district <- factor(mini_htx2$district)

plot(mini_htx2$lat, mini_htx2$lng, col = mini_htx2$district, panel.first = grid(8,8),  xlim = c(29.5, 30.15), ylim = c(-96, -95), main = "Latitude & Longitude of Stops by District", xlab = "Latitude", ylab = "Longitude", pch=1)
legend("bottomright", legend=levels(mini_htx2$district), cex = 0.62, pch=1, col=unique(mini_htx2$district), title = "District")
```

Next, we want to explore the spatial features of our data. We have both the latitude and the longitude of each stop, along with the district and beat. We knew that beat meant a motorized police unit that patrols a specific territory, but “district” in this context is not as clear, as it could mean multiple things. We investigated this by creating a scatterplot of the latitude versus the longitude of stops and colored it by district. We see very clear spatial grouping for each district, which means districts are not a police-defined feature but a location feature. From this plot, we can ascertain that district refers to Houston's subdistricts. [As seen here: https://www.houstontx.gov/police/pdfs/hpd_beat_map.pdf]

```{r, echo=FALSE}
#mini_htx2 <- mini_htx[seq(1,666666,6),]
mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
mini_htx3 <- mini_htx2[, c("date", "diff")]
mini_htx3 <- na.omit(mini_htx3)
mini_htx3 <- mini_htx3[mini_htx3$diff != 693,]

plot2 <- ggplot(mini_htx3, aes(date, diff)) + geom_point(size=.5) +
  labs(subtitle="Frequency of Citations for Different Levels of Speeding", 
       y="MPH Speeding", 
       x="Year", 
       title="MPH over Limit for Speeding Citations 2014-2020")

ggMarginal(plot2, type = "histogram", fill="slateblue", margins = 'y')
#ggMarginal(plot2, type = "boxplot", fill="slateblue", margins = 'y')
#ggMarginal(plot2, color ="slateblue", margins = 'y')
```

This plot investigates citations that are given for speeding. Namely, by how much were people speeding over the limit to have been issued a citation. The y axis is mph over the speed limit (as recorded on the citation). The x axis is the year/general time from the citation occurred in, and there does seem to be some evidence of temporal trends. There is a vertical line before 2018 where it seems like there were less tickets given overall, which could be a number of things that we could look into. It could be something about how the data was collected, if there was some data lost at any point, or maybe HPD deprioritized patrolling for speeding during that time period. There is no way to tell from just looking at the graph. Another interesting feature is that there seems to be less citations given for speeding less than 10mph over the speed limit 2020 and onward. It is also interesting that there are speeding citations given for going as little as 1-5 mph over the speed limit. The most frequent citations were given around 10-15 mph over the speed limit, as seen by the histogram of frequent cited mph over speed limit on the left margin.

```{r, echo=FALSE}
mini_htx4 <- na.omit(mini_htx2[,c("subject_race", "date", "diff", "subject_sex")])
mini_htx4 <- mini_htx4[mini_htx4$diff != 693,]

ggplot(mini_htx4, aes(fill=factor(subject_sex),subject_race, diff)) + geom_violin() + labs(title="MPH over Speed Limit for Citation", x="Subject Race",y="MPH Speeding")+guides(fill=guide_legend(title='Subject Sex'))

```

Moving forward, we chose to visualize MPH over Speed Limit for Citation by Subject race and sex. We visualized this with violin plots, which demonstrate the density and distribution of the data. We observed that each race except for "unknown" had a fairly similar distribution of speeding amounts, with the greatest peak around 10 MPH. Distributions of speeding amounts for each sex within each race were also very similar, although males of every race reached a greater maximum speeding amount than females.

```{r, echo=FALSE}
#x axis citation number, y axis is district, color is gender

library(ggthemes)
options(scipen = 999)  # turns of scientific notations like 1e+40

test <- table(mini_htx$subject_sex, mini_htx$district)
new_df <- data.frame(matrix(ncol = 3, nrow = 0))
for (row in rownames(test)){
  for (col in colnames(test)){
    #row in new dataframe
    #(sex, district, value)
    new_df <- rbind(new_df, c(row, col, test[row, col]))
  }
}
colnames(new_df) <- c("sex", "district", "count")
new_df$count <- as.numeric(new_df$count)
new_df$district <- as.factor(new_df$district)

mens <- new_df[new_df$sex == "male",]
mens$count <- (mens$count * -1)
new_df <- rbind(new_df[new_df$sex == "female",], mens)
new_df$district <- ordered(new_df$district, seq(1,24))

brks <- seq(-50000, 30000, 10000)
lbls = paste0(as.character(c(seq(500, 0, -100), seq(100, 300, 100))), "k")
#lbls = paste0(as.character(c(seq(15, 0, -5), seq(5, 15, 5))), "m")

# Plot
library(ggfortify)
ggplot(new_df, aes(x = district, y = count, fill = sex)) +   # Fill column
                              geom_bar(stat = "identity", width = .7) +   # draw the bars
                              scale_y_continuous(breaks = brks, labels = lbls) + 
                              coord_flip() +  # Flip axes
                              labs(title="Citation by Sex by District - Pyramid", y = "Citation Count", x = "District") +
                              theme_tufte() +  # Tufte theme from ggfortify
                              theme(plot.title = element_text(hjust = .5), 
                                    axis.ticks = element_blank()) +   # Centre plot title
                              scale_fill_brewer(palette = "Dark2")  # Color palette

```

We created a "population pyramid" of citation count by district and by sex. The x-axis is the amount of citations in thousands, and the y-axis is the police district in Houston. This plot reveals interesting aspects of our dataset regarding the proportion of citations by gender. There are more citations for men than women for every single districts. This could be because men are more likely to speed, or that women speed the same but are not given citations as often. The core reason for the dependency is unknown, but the unequal count by sex is clear to see. Also, as found before in other graphs, district 21 and 23 have the least amount of citations as they are both airports (IAH and Hobby), and people are simply unable to speed in the same capacity in airport roads as they can on a highway. District 2 has the most citations, and this corresponds to the Houston's Greater Heights district, roughly.

## Questions 
Investigating aspects of our data with supplemental data as well.

```{r Establish Connection, include=FALSE}
# Connect to database
con <- dbConnect(RSQLite::SQLite(), dbname = "policedb")

```
### 1- How does the racial breakdown of citations compare to ther racial breakdown of Houston?

From our main dataset, we visualized the breakdown of citations per race. Our group wanted to see how the breakdown of citations per race compared to the breakdown of race in Houston.      
```{r, echo=FALSE}
citations <- dbGetQuery(con, "SELECT subject_race AS race, 1.0*COUNT(*)/SUM(COUNT(*)) OVER() AS citation_prop
                        FROM 'mini_htx2'
                        WHERE subject_race IS NOT NULL
                        GROUP BY subject_race
                        ORDER BY race")
population <- dbGetQuery(con, "SELECT CASE
                                      WHEN Fact = 'White alone, percent' THEN 'white'
                                      WHEN Fact = 'Black or African American alone, percent' THEN 'black'
                                      WHEN Fact = 'American Indian and Alaska Native alone' THEN 'american indian/alaska native'
                                      WHEN Fact IN ('Asian alone, percent', 'Native Hawaiian and Other Pacific Islander alone, percent') THEN 'asian/pacific islander'
                                      WHEN Fact = 'Two or More Races, percent' THEN 'other'
                                      ELSE Fact
                                  END AS race,
                                  SUM(`Houston city, Texas`/100) AS population_prop
                                  FROM
                                    (SELECT *
                                     FROM census
                                     WHERE Fact IN ('White alone, percent', 'Black or African American alone, percent', 'American Indian and Alaska Native alone', 'Asian alone, percent', 'Native Hawaiian and Other Pacific Islander alone, percent', 'Two or More Races, percent')) AS subquery
                                  GROUP BY race
                         ORDER BY race")
htx <- inner_join(citations, population, by = "race")
#htx
knitr::kable(htx)

```

From this table, we can see that the population proportion of White people is lower than their citation proportion, the population, the population proportion of Black people is lower than their citation proportion, and the population proportion of Asian/Pacific Islander people is higher than the citation proportion.The 'other' population proportion is significantly higher than the citation proportion, but this may be because we signified mixed-race individuals as others, while for citation purposes, people giving citations may mark them as a single race.

### 2- What is the race breakdown of each type of citation?

Looking at our data, our team noticed that there were 5 main types of violations: speeding, invalid license, failure to establish financial responsibility, failure to wear a seat belt, and running a stop sign/red light. We wanted to analyze the racial breakdown of each type of citation in order to see if any racial group disproportionately received any type of citation.

Speeding Citations by Race:
```{r q1, echo=FALSE}
query <- "
SELECT subject_race, COUNT(*) AS count
FROM mini_htx2
WHERE violation LIKE '%SPEEDING%'
GROUP BY subject_race
ORDER BY count DESC"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)

```

Invalid License Citations by Race:
```{r q1.2, echo=FALSE}
dbClearResult(res)
query <- "
SELECT subject_race, COUNT(*) AS count
FROM mini_htx2
WHERE violation LIKE '%LICENSE%'
GROUP BY subject_race
ORDER BY count DESC"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)
```

Failure to Establish Financial Responsibility Citations by Race:
```{r q1.3, echo=FALSE}
# Failure to establish Financial Responsibility Citations
dbClearResult(res)
query <- "
SELECT subject_race, COUNT(*) AS count
FROM mini_htx2
WHERE violation LIKE '%FINANCIAL RESPONSIBILITY%'
GROUP BY subject_race
ORDER BY count DESC"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)
```

Seat Belt Citations by Race:
```{r q1.4, echo=FALSE}
# seat belt citations
dbClearResult(res)
query <- "
SELECT subject_race, COUNT(*) AS count
FROM mini_htx2
WHERE violation LIKE '%SEAT BELT%'
GROUP BY subject_race
ORDER BY count DESC"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)
```

Running a Red Light/Stop Sign Citations by Race:
```{r q1.5, echo=FALSE}
# Running a stop light/red light
dbClearResult(res)
query <- "
SELECT subject_race, COUNT(*) AS count
FROM mini_htx2
WHERE violation LIKE '%RED LIGHT%' OR violation LIKE '%STOP SIGN%'
GROUP BY subject_race
ORDER BY count DESC"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)
dbClearResult(res)
```

For nearly all types of citations, when considering subjects with defined races, white individuals received the most citations, followed by black and AAPI individuals; this aligns with the racial breakdown of the city of Houston.      
However, black individuals received the most citations for failure to establish financial responsibility, which refers to the inability of the subject to provide proof of insurance. Typically, this citation should only be issued given that the subject has committed some other infraction that necessitates police interaction and request for proof of insurance. So, suspicion may be raised around citations that record failure to establish financial responsibility as the sole infraction, as the officer has not recorded any indication of why the individual was pulled over in the first place. Citations that only reference failure to establish financial responsibility may be useful in identifying possible racial bias, as the officer may have pulled over the individual based on their appearance/race, since they did not indicate any other offense on the citation. This led our team to investigate what proportion of citations for individuals of all races resulted solely from failure to establish financial responsibility.

### 3- What proportion of citations for individuals of each race resulted solely from failure to establish financial responsibility?
```{r, echo=FALSE, include=FALSE}
dropquery <- "
DROP VIEW IF EXISTS race_totals;
DROP VIEW IF EXISTS finresp_race_totals;"

query1 <- "
CREATE VIEW IF NOT EXISTS race_totals AS
SELECT subject_race, COUNT(*) AS total
FROM mini_htx2
GROUP BY subject_race;"

query2 <- "
CREATE VIEW IF NOT EXISTS finresp_race_totals AS
SELECT subject_race, COUNT(*) AS fin_fail
FROM mini_htx2
WHERE violation='FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY'
GROUP BY subject_race;"

dbSendQuery(con, query1)
dbSendQuery(con, query2)
#dbExecute(con, dropquery)
```

```{r, echo=FALSE}
query <- "
SELECT race_totals.subject_race, CAST(fin_fail AS real)/total AS fin_fail_percentage
FROM race_totals
JOIN finresp_race_totals ON race_totals.subject_race = finresp_race_totals.subject_race
GROUP BY race_totals.subject_race;"
res <- dbSendQuery(con, query)
out <- dbFetch(res)
knitr::kable(out)
dbClearResult(res)

# different results from dplyr because dplyr does not filter out N/As from race totals, so race totals were higher for every race

```

Of all recorded races, the proportion of total citations that mention failure to establish financial responsibility as the sole violation is the highest for black individuals. This proportion is about 0.02 higher for black individuals, while all other races share a similar proportion around 0.02.

### 4- What is the average speeding amount that results in a citation? For each racial group? For each gender?

Next, we wanted to investigate if different races and sexes received speeding citations equally, or if certain groups were given citations with different frequencies or for different severity of speeding.

All People (baseline):
```{r, echo=FALSE}
## All people
query <- "
SELECT AVG(speed-posted_speed) AS avg_diff, STDEV(speed-posted_speed) as std_diff
FROM mini_htx2
WHERE posted_speed >= 0"
res <- dbSendQuery(con, query)
df1 <- dbFetch(res)
dbClearResult(res)
knitr::kable(df1)

```

By Race:
```{r, echo=FALSE}
## By Race
query <- "
SELECT subject_race, COUNT(*) AS count, AVG(speed-posted_speed) AS avg_diff, STDEV(speed-posted_speed) AS std_diff
FROM mini_htx2
WHERE posted_speed >= 0 
GROUP BY subject_race
ORDER BY avg_diff ASC"
res <- dbSendQuery(con, query)
df2 <- dbFetch(res)
dbClearResult(res)
knitr::kable(df2)
```

One thing to note is the 'unknown' category. Police officers fill out the reports without asking the subjects they pulled over about their race or ethnicity. All of these reported races are from the perspective of the police officers and therefore it is difficult to know the true breakdown.

Here, the unknown could honestly be evidence of bias. We see that for people the officers are unsure about their race (and that they do not ask to clarify), they are given citations for slower speeds (instead of being let off with a warning).

By Gender:
```{r, echo=FALSE}
## By Gender
query <- "
SELECT subject_sex, COUNT(*) AS count, AVG(speed-posted_speed) AS avg_diff, STDEV(speed-posted_speed) as std_diff
FROM mini_htx2
WHERE posted_speed >= 0 
GROUP BY subject_sex
ORDER BY avg_diff ASC"
res <- dbSendQuery(con, query)
df3 <- dbFetch(res)
dbClearResult(res)
knitr::kable(df3)
```

Men and women are ticketed for roughly the same speeds, bit there are FAR more men ticketed than women in our dataset. This is not equal to the proportion of men and women in Houston, so this is either evidence that men speed more often than women, or that officers are more likely to give men citations and let women off with a warning, or that women are simply caught speeding less often, but speed just as frequently.

By Race AND Gender:
```{r, echo=FALSE}
## By Race AND Gender
query <- "
SELECT subject_race, COUNT(*) AS n, AVG(speed-posted_speed) AS avg_diff, STDEV(speed-posted_speed) as std_diff
FROM mini_htx2
WHERE posted_speed >= 0 
GROUP BY subject_race, subject_sex
ORDER BY avg_diff ASC"
res <- dbSendQuery(con, query)
df4 <- dbFetch(res)
dbClearResult(res)
knitr::kable(df4)
```

Unknown race females have the lowest average amount speeding of 11.8mph and median of 10mph, the lowest of any group.

Since all of the race and sex data is reported by the officers, if there were bias, it could be easily hidden in the "unknowns" or "NAs" in the data. We are able to see that men get ticketed with a much greater frequency than women, but between race there are not many conclusions we can make. There is one exception, however. For most races it is true that men are ticketed two times as much as women of the same race, with the exception being Black people. Black women are ticketed almost as much as black men (24.4k and 29.9k, respectively). This could be evidence of bias against black women.

### 5- What time of day do citation happen?
Our group decided to look at the frequency of citations by hour of day.

```{r Problem 4, echo=FALSE, results='hide'}
hour_count <- dbGetQuery(con, "SELECT strftime('%H', time, 'unixepoch') AS hour, COUNT(*) AS count
                 FROM mini_htx2
                 WHERE time IS NOT NULL
                 GROUP BY hour
                 ORDER BY count DESC")
#print(paste("The lowest number of citations occur around", hour_count$hour[24], "hr"))
#print(paste("The highest number of citations occur around", hour_count$hour[1], "hr"))
#print(paste("The highest number of citations occur at", paste(hour_count$hour[1:5], collapse = ", "), "hr"))

```

*The lowest number of citations occur around `r hour_count$hour[24]` hr.*

*The highest number of citations occur around `r hour_count$hour[1]` hr.*

*The highest number of citations occur at `r paste(hour_count$hour[1:5], collapse = ", ")` hrs.*

We can see that the highest number of numbers of citations occur, in descending order, at 8am, 3pm, 2pm, 4pm an 1pm. Most of these times are in the afternoon, except for the 8am time. There might be a large number of citations at around 8am because this is the time that most people are travelling to work in the morning. Because of the morning rush and the need to get to work on time, people may be more likely to speed/commit citation-able offences at this time.

The rest of the times where the highest number of citations are given are in the afternoon, between 1pm - 4pm. This may also be attributed to rush hour traffic again, and things like school dismissal -- police may be more vigilant around school zones, while there may be more people on the roads due to schools getting out.

The smallest number of citations occur around 4am. This may be due to a lack of people on the road, because it is very late at night/early in the morning.

### 6- What is the number of citations per square mile for each beat?

Next, our team aimed to understand the geographic distribution of citations by calculating the number of citations per square mile within each beat. To aid with our investigation, we found the median, standard deviation, and interquartile range of the number of citations per square mile. We also looked into the 5 beats with the greatest amount of citations per square mile and the 5 beats with the least amount of citations per square mile.

```{r, echo=FALSE}
# join the two dataframes on beats

citations_in_beat <- mini_htx %>% inner_join(beats_data, by="Beats") %>%
  group_by(Beats) %>%
  summarise(n_individuals = n())

citations_in_beat <- as.data.frame(citations_in_beat)

cit_beat_area <- citations_in_beat %>% 
  inner_join(beats_data[,c("Beats", "Area_sq_mi")], by="Beats")

cit_beat_area$Citations_per_sqm <- cit_beat_area$n_individuals / cit_beat_area$Area_sq_mi
csqm <- cit_beat_area[,c("Beats","Citations_per_sqm")]
csqm2 <- csqm[order(-csqm$Citations_per_sqm),]
knitr::kable(csqm)

```

**Statistics on number of citations/square mile:**
```{r, echo=FALSE}
cqm_summary <- data.frame(matrix(ncol=3,nrow=1))
colnames(cqm_summary) <- c("median","standard deviation", "IQR")
cqm_summary[,1] <- csqm %>% summarise(median(Citations_per_sqm))
cqm_summary[,2] <- csqm %>% summarise(sd(Citations_per_sqm))
cqm_summary[,3] <- csqm %>% summarise(IQR(Citations_per_sqm))

knitr::kable(cqm_summary)
```

**Top 5 Beats ranked by citations/square mile:**
```{r, echo=FALSE, include=FALSE}
# join the two dataframes on beats
dropquery <- "
DROP VIEW IF EXISTS citations_in_beat;
DROP VIEW IF EXISTS cit_beat_area;"

query1 <- "
CREATE VIEW IF NOT EXISTS citations_in_beat AS
SELECT Beats AS beats_c, COUNT(*) as count
FROM mini_htx2 JOIN beats_data ON mini_htx2.beat = beats_data.Beats
GROUP BY Beats;"

dbExecute(con, dropquery)
dbSendQuery(con, query1)
```

```{r, echo=FALSE}
query3 <- "
SELECT citations_in_beat.beats_c, CAST(citations_in_beat.count AS real)/beats_data.Area_sq_mi AS citations_per_sqm
FROM citations_in_beat JOIN beats_data ON citations_in_beat.beats_c=beats_data.Beats
ORDER BY citations_per_sqm DESC LIMIT 5;"

res <- dbSendQuery(con, query3)
out <- dbFetch(res)
dbClearResult(res)
knitr::kable(out)
```
Upon comparison to a map of Houston's police beats, we saw that all five of these beats are adjacent, small, and located in the heart of Houston. Additionally, we noted that both beats 1A10 and 2A40 are very small in terms of square mileage, but contain two and one police stations, respectively; proximity to police stations could therefore explain the extremely high citations/square mile values for these beats.

**Bottom 5 Beats ranked by citations/square mile:**
```{r, echo=FALSE}

query2 <- "
SELECT citations_in_beat.beats_c, CAST(citations_in_beat.count AS real)/beats_data.Area_sq_mi AS citations_per_sqm
FROM citations_in_beat JOIN beats_data ON citations_in_beat.beats_c=beats_data.Beats
ORDER BY citations_per_sqm ASC LIMIT 5;"

out <- dbGetQuery(con, query2)
knitr::kable(out)
```
Referencing a map of Houston's police beats, our team found that these five beats were located in Houston's suburbs and are all relatively large in terms of square mileage, which explains their low citations/square mile values.

Map Reference: https://www.houstontx.gov/police/pdfs/hpd_beat_map.pdf

```{r Break Connection, echo=FALSE}

dbDisconnect(con)

```

## Conclusion

In our investigation, we scanned for evidence of racial bias but did not found statistically significant instances of bias. For example, each type of citation had a racial breakdown similar to the overall population and citation breakdown. We also found that beats with the greatest proportion of citations to square mileage are those located in the heart of Houston that are densely populated with police stations, while the beats with the lowest ratio were suburbs with large square mileage at a greater distance from stations. There are interesting observations like black men and women are given speeding citations approximately at the same rates, while all other races men are given citations almost 2x as much. However, it is a stretch to confidently draw the conclusion of bias from a one off finding like that. We are excited to hopefully receive more data from HPD via FOIA request to bolster this analysis from beyond what is available publicly.

### 2023-04-13 ui.R
shiny plot

@@ -1,0 +1,35 @@
+#
+# This is the user-interface definition of a Shiny web application. You can
+# run the application by clicking 'Run App' above.
+#
+# Find out more about building applications with Shiny here:
+#
+#    http://shiny.rstudio.com/
+#
+
+library(shiny)
+
+# Define UI for application that draws a histogram
+shinyUI(fluidPage(
+
+    # Application title
+    titlePanel("Citation Breakdown by District"),
+
+    # Sidebar with a slider input for number of bins
+    sidebarLayout(
+        sidebarPanel(
+            checkboxGroupInput(inputId = "districts",
+                        "Districts to show:",
+                        c("1", "2", "3", "4", "5",
+                          "6", "7", "8", "9", "10",
+                          "11", "12", "13", "14", "15",
+                          "16", "17", "18", "19", "20",
+                          "21", "22", "23", "24"))
+        ),
+
+        # Show a plot of the generated distribution
+        mainPanel(
+            plotOutput("killerplot")
+        )
+    )
+))

### 2023-04-13 server.R
server update that will not work tehe

@@ -233,7 +233,9 @@ shinyServer(function(input, output) {
           
           # make circle
           count <- pull(citations_htx_circle[citations_htx_circle['subject_race']==races[j-1] &
-                                               citations_htx_circle['date']==years[i],"n_individuals"])[1]
+                                               citations_htx_circle['date']==years[i] &
+                                               citations_htx_circle['district'] == as.numeric(input$districts),
+                                             "n_individuals"])[1]
           #max_count <- sum(citations_htx_circle$n_individuals)
           max_count <- max(citations_htx_circle$n_individuals)
           draw_circle(count, max_count)

### 2023-04-13 Presentation.Rmd
created presentation rmd

---
title: "Presentation"
author: "Virginia Baskin"
date: "4/13/2023"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Shiny Presentation

This R Markdown presentation is made interactive using Shiny. The viewers of the presentation can change the assumptions underlying what's presented and see the results immediately.

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Interactive Plot

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),

  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")

  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## R Output

```{r cars}
summary(cars)
```

## Data 
- Problem and Importance

## Data Introduction 
- bullet points 
- bullet points 
- bullet points 
- dude our data was all categorical so when i tell u it was difficult to plot

## Supplementary Datasets

## Exploration

--- 
plots
- plot of citations by race over time
- plot of frequency of citation versus hour in day
- mph over speed limit by year

---
plot

---

```{r}
#source("plots/plot.r", local = TRUE, print.eval = TRUE)
```

--- 
## Killer Plot

## Shiny

---

### 2023-04-13 server.R
shiny plot fixes

@@ -6,143 +6,7 @@
 #
 #    http://shiny.rstudio.com/
 #
-
-library(shiny)
-
 # import packages
-library(RSQLite)
-library(dplyr)
-library(stringr)
-library(grid)
-library(readr)
-library(shiny)
-options(dplyr.summarise.inform = FALSE)
-
-# import data
-# full_htx <- read_csv("/stat405final/Datasets/tx_houston_2023_01_26.csv")
-census_csv <- read_csv("Datasets/Census Data Houston.csv")
-mini_htx <- read_csv("Datasets/mini_htx.csv")
-# mini_htx <- full_htx[seq(1,2000000,3),]
-mini_htx <- mini_htx %>% rename("Beats" = "beat")
-#beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
-#htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")
-
-# subset the data to only contain columns of interest
-citations_htx <- mini_htx[,c("subject_race", "subject_sex", "district", "date", "violation")]
-
-# subset the data to only contain rows of violation interest
-citations_htx <- citations_htx %>% 
-  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)|
-           grepl("SEAT BELT", violation)|grepl("FINANCIAL RESPONSIBILITY", violation)|
-           grepl("LICENSE", violation)|grepl("SPEEDING", violation)) %>%
-  mutate(date = format(date, format = "%Y"))
-
-#create the circles dataframe which counts individuals in each race, year, district pairing
-citations_htx_circle <- citations_htx %>%
-  group_by(subject_race, date, district) %>%
-  summarise(n_individuals = n())
-citations_htx_circle
-
-# create the bars dataframe which counts the number of individuals in each race, year, district, violation, pairing
-
-# create a dataframe with all possible combinations of race, date, sex, and violation
-citations_htx_bar <- citations_htx %>%
-  mutate(red_stop= if_else(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation), 1, 0)) %>%
-  mutate(seat_belt= if_else(grepl("SEAT BELT", violation), 1, 0)) %>%
-  mutate(fin_resp= if_else(grepl("FINANCIAL RESPONSIBILITY", violation), 1, 0))%>%
[…]

### 2023-04-13 shiny_plot_setup.R

@@ -1,0 +1,135 @@
+library(RSQLite)
+library(dplyr)
+library(stringr)
+library(grid)
+library(readr)
+options(dplyr.summarise.inform = FALSE)
+
+# import data
+# full_htx <- read_csv("/stat405final/Datasets/tx_houston_2023_01_26.csv")
+census_csv <- read_csv("Datasets/Census Data Houston.csv")
+mini_htx <- read_csv("Datasets/mini_htx.csv")
+# mini_htx <- full_htx[seq(1,2000000,3),]
+mini_htx <- mini_htx %>% rename("Beats" = "beat")
+#beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
+#htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")
+
+# subset the data to only contain columns of interest
+citations_htx <- mini_htx[,c("subject_race", "subject_sex", "district", "date", "violation")]
+
+# subset the data to only contain rows of violation interest
+citations_htx <- citations_htx %>% 
+  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)|
+           grepl("SEAT BELT", violation)|grepl("FINANCIAL RESPONSIBILITY", violation)|
+           grepl("LICENSE", violation)|grepl("SPEEDING", violation)) %>%
+  mutate(date = format(date, format = "%Y"))
+
+#create the circles dataframe which counts individuals in each race, year, district pairing
+citations_htx_circle <- citations_htx %>%
+  group_by(subject_race, date, district) %>%
+  filter_at(vars("district"), all_vars(!is.na(.)))%>%
+  summarise(n_individuals = n()) 
+
+# create the bars dataframe which counts the number of individuals in each race, year, district, violation, pairing
+
+# create a dataframe with all possible combinations of race, date, sex, and violation
+citations_htx_bar <- citations_htx %>%
+  mutate(red_stop= if_else(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation), 1, 0)) %>%
+  mutate(seat_belt= if_else(grepl("SEAT BELT", violation), 1, 0)) %>%
+  mutate(fin_resp= if_else(grepl("FINANCIAL RESPONSIBILITY", violation), 1, 0))%>%
[…]

### 2023-04-13 ui.R

@@ -8,6 +8,7 @@
 #
 
 library(shiny)
+source("shiny_plot_setup.R")
 
 # Define UI for application that draws a histogram
 shinyUI(fluidPage(
@@ -18,13 +19,9 @@ shinyUI(fluidPage(
     # Sidebar with a slider input for number of bins
     sidebarLayout(
         sidebarPanel(
-            checkboxGroupInput(inputId = "districts",
-                        "Districts to show:",
-                        c("1", "2", "3", "4", "5",
-                          "6", "7", "8", "9", "10",
-                          "11", "12", "13", "14", "15",
-                          "16", "17", "18", "19", "20",
-                          "21", "22", "23", "24"))
+          sliderInput("districts", "District:",
+                      min = 1, max = 24,
+                      value = 1, step = 1)
         ),
 
         # Show a plot of the generated distribution

### 2023-04-16 shiny_plot_setup.R
shiny plot fixes

@@ -27,7 +27,7 @@ citations_htx <- citations_htx %>%
 #create the circles dataframe which counts individuals in each race, year, district pairing
 citations_htx_circle <- citations_htx %>%
   group_by(subject_race, date, district) %>%
-  filter_at(vars("district"), all_vars(!is.na(.)))%>%
+  na.omit()%>%
   summarise(n_individuals = n()) 
 
 # create the bars dataframe which counts the number of individuals in each race, year, district, violation, pairing
@@ -70,9 +70,9 @@ barplot <- function(red_stop_m, red_stop_f, seat_belt_m, seat_belt_f, fin_resp_m
 }
 
 bar_run <- function(plot_race,plot_year, districts){
-  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year), ]
-  new_df <- new_df[which(citations_htx_bar$subject_race == plot_race),]
-  new_df <- new_df[which(citations_htx_bar$district %in% districts),]
+  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year &
+                                      citations_htx_bar$subject_race==plot_race &
+                                      citations_htx_bar$district==districts), ]
   male_df <- new_df[which(new_df$subject_sex == "male"),]
   female_df <- new_df[which(new_df$subject_sex == "female"),]
   red_stop_m <- sum(male_df$red_stop == 1)

### 2023-04-16 draft4.Rmd
new plot scripts

@@
 
-legend("topright", legend=c("black", "white", "aapi", "unknown"), col=c("blue", "red", "green", "orange"), pch = 20, cex = 0.7, title = "Race")
+  other <- race_date[race_date$subject_race=="other",]
+  other$year <- unlist(lapply(other$date, year_only))
+  points(sort(unique(other$year)), as.vector(table(as.factor(other$year))))
+  lines(sort(unique(other$year)), as.vector(table((as.factor(other$year)))), 
+        type = "l", col = "purple")
+
+legend("topright", legend=c("black", "white", "aapi", "unknown", "other"), col=c("blue", "red", "green", "orange", "purple"), pch = 20, cex = 0.7, title = "Race")
 
@@
 #mini_htx2 <- mini_htx[seq(1,666666,6),]
+
 mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)

### 2023-04-16 shiny_plot_setup.R

@@ -3,6 +3,8 @@ library(dplyr)
 library(stringr)
 library(grid)
 library(readr)
+library(ggfortify)
+library(ggthemes)
 options(dplyr.summarise.inform = FALSE)
 
 # import data

### 2023-04-16 subscripts/citation_pyramid.R

@@ -1,0 +1,42 @@
+#citation pyramid 
+source("shiny_plot_setup.R")
+
+#x axis citation number, y axis is district, color is gender
+pyramid_cites <- function(){
+  options(scipen = 999)  # turns of scientific notations like 1e+40
+  
+  test <- table(mini_htx$subject_sex, mini_htx$district)
+  new_df <- data.frame(matrix(ncol = 3, nrow = 0))
+  for (row in rownames(test)){
+    for (col in colnames(test)){
+      #row in new dataframe
+      #(sex, district, value)
+      new_df <- rbind(new_df, c(row, col, test[row, col]))
+    }
+  }
+  colnames(new_df) <- c("sex", "district", "count")
+  new_df$count <- as.numeric(new_df$count)
+  new_df$district <- as.factor(new_df$district)
+  
+  mens <- new_df[new_df$sex == "male",]
+  mens$count <- (mens$count * -1)
+  new_df <- rbind(new_df[new_df$sex == "female",], mens)
+  new_df$district <- ordered(new_df$district, seq(1,24))
+  
+  brks <- seq(-50000, 30000, 10000)
+  lbls = paste0(as.character(c(seq(500, 0, -100), seq(100, 300, 100))), "k")
+  #lbls = paste0(as.character(c(seq(15, 0, -5), seq(5, 15, 5))), "m")
+  
+  # Plot
+  
+  ggplot(new_df, aes(x = district, y = count, fill = sex)) +   # Fill column
+    geom_bar(stat = "identity", width = .7) +   # draw the bars
+    scale_y_continuous(breaks = brks, labels = lbls) + 
+    coord_flip() +  # Flip axes
+    labs(title="Citation by Sex by District - Pyramid", y = "Citation Count", x = "District") +
+    theme_tufte() +  # Tufte theme from ggfortify
+    theme(plot.title = element_text(hjust = .5), 
+          axis.ticks = element_blank()) +   # Centre plot title
+    scale_fill_brewer(palette = "Dark2")  # Color palette
+  
+}

### 2023-04-16 subscripts/district_scatter.R

@@ -1,0 +1,16 @@
+source("shiny_plot_setup.R")
+#district scatter
+
+#Scatter plot longitude vs latitude by district
+# Verifying the hubs for citation and locations of districts
+
+district_loc <- function(){ 
+  mini_htx2 <- mini_htx[seq(1,666666,30),]
+  mini_htx2$district <- factor(mini_htx2$district)
+  
+  plot(mini_htx2$lat, mini_htx2$lng, col = mini_htx2$district, panel.first = grid(8,8),  
+       xlim = c(29.5, 30.15), ylim = c(-96, -95), main = "Latitude & Longitude of Stops by District", 
+       xlab = "Latitude", ylab = "Longitude", pch=1)
+  legend("bottomright", legend=levels(mini_htx2$district), cex = 0.62, pch=1, 
+         col=unique(mini_htx2$district), title = "District")
+}

### 2023-04-16 subscripts/line_race_year.R

@@ -1,0 +1,50 @@
+#citation count by race by year line graph
+#Race Line Graph by Year: line_ry()
+
+source("shiny_plot_setup.R")
+
+race_date <- mini_htx[,c("date", "subject_race")] # race and date
+race_date <- na.omit(race_date) # omit na Values
+
+year_only <- function(fulldate){
+  year <- substring(fulldate, 1, 4) # only keep year
+}
+
+line_ry <- function(){ 
+  black <- race_date[race_date$subject_race=="black",]
+  black$year <- unlist(lapply(black$date, year_only))
+  plot(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), 
+       ylim = c(0, 60000), main = "Citations by Race for Each Year", 
+       xlab="Year", ylab = "Count") #200000 when all
+  lines(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), 
+        type = "l", col = "blue")
+  
+  white <- race_date[race_date$subject_race=="white",]
+  white$year <- unlist(lapply(white$date, year_only))
+  points(sort(unique(white$year)), as.vector(table(as.factor(white$year))))
+  lines(sort(unique(white$year)), as.vector(table((as.factor(white$year)))), 
+        type = "l", col = "red")
+  
+  aapi <- race_date[race_date$subject_race=="asian/pacific islander",]
+  aapi$year <- unlist(lapply(aapi$date, year_only))
+  points(sort(unique(aapi$year)), as.vector(table(as.factor(aapi$year))))
+  lines(sort(unique(aapi$year)), as.vector(table((as.factor(aapi$year)))), 
+        type = "l", col = "green")
+  
+  
+  unknown <- race_date[race_date$subject_race=="unknown",]
+  unknown$year <- unlist(lapply(unknown$date, year_only))
+  points(sort(unique(unknown$year)), as.vector(table(as.factor(unknown$year))))
+  lines(sort(unique(unknown$year)), as.vector(table((as.factor(unknown$year)))), 
+        type = "l", col = "orange")
+  
+  other <- race_date[race_date$subject_race=="other",]
+  other$year <- unlist(lapply(other$date, year_only))
+  points(sort(unique(other$year)), as.vector(table(as.factor(other$year))))
+  lines(sort(unique(other$year)), as.vector(table((as.factor(other$year)))), 
+        type = "l", col = "purple")
+  
[…]

### 2023-04-16 subscripts/line_timeofcites.R

@@ -1,0 +1,11 @@
+source("shiny_plot_setup.R")
+
+#ADD COLOR TO THIS PLOT 
+# Observe citations over time and any time-associate patterns
+time_line <- function(){ 
+  time <- as.POSIXct(mini_htx$time, format = "%H:%M:%S")
+  hour <- as.numeric(format(time, "%H"))
+  count <- data.frame(table(na.omit(hour)))
+  plot(count$Var1, count$Freq, type = "n", xlab = "Hour", ylab = "Frequency of Citations", main = "Frequency of Citations vs Hour in Day")
+  lines(count$Var1, count$Freq, type = "l", lwd = 2)
+}

### 2023-04-16 subscripts/speed_over_freq.R

@@ -1,0 +1,22 @@
+source("shiny_plot_setup.R")
+
+#frequency of different levels
+speed_over_freq <- function(){
+
+  mini_htx2 <- mini_htx[seq(1,666666,30),]
+  mini_htx2$district <- factor(mini_htx2$district)
+  mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
+  mini_htx3 <- mini_htx2[, c("date", "diff")]
+  mini_htx3 <- na.omit(mini_htx3)
+  mini_htx3 <- mini_htx3[mini_htx3$diff != 693,]
+  
+  plot2 <- ggplot(mini_htx3, aes(date, diff)) + geom_point(size=.5) +
+    labs(subtitle="Frequency of Citations for Different Levels of Speeding", 
+         y="MPH Speeding", 
+         x="Year", 
+         title="MPH over Limit for Speeding Citations 2014-2020")
+  
+  ggMarginal(plot2, type = "histogram", fill="slateblue", margins = 'y')
+  #ggMarginal(plot2, type = "boxplot", fill="slateblue", margins = 'y')
+  #ggMarginal(plot2, color ="slateblue", margins = 'y')
+}

### 2023-04-16 subscripts/violin_speed.R

@@ -1,0 +1,11 @@
+source("shiny_plot_setup.R")
+
+violin_speed <- function(){
+  mini_htx4 <- na.omit(mini_htx2[,c("subject_race", "date", "diff", "subject_sex")])
+  mini_htx4 <- mini_htx4[mini_htx4$diff != 693,]
+  
+  ggplot(mini_htx4, aes(fill=factor(subject_sex),subject_race, diff)) + 
+    geom_violin() + 
+    labs(title="MPH over Speed Limit for Citation", x="Subject Race",y="MPH Speeding")+
+    guides(fill=guide_legend(title='Subject Sex'))
+}

### 2023-04-16 subscripts/waffle.R

@@ -1,0 +1,26 @@
+
+
+source("shiny_plot_setup.R")
+
+#Race waffle plot based on citation counts  
+df <- data.frame(table(na.omit(mini_htx$subject_race)))
+total_freq <- sum(df$Freq)
+df$prop <- df$Freq/total_freq
+num_squares <- round(df$prop*100)
+
+waffle <- function(){  
+  df_grid <- data.frame(
+    x = rep(1:10, each = 10),
+    y = rep(1:10, times = 10)
+  )
+  num_squares <- c(3, 36, 1, 3, 57) #manually add one more square because the rounding resulted in only 99 
+  #num_squares <- round(df$prop*100)
+  
+  df_grid$category <- rep(df$Var1, num_squares)
+  ggplot(df_grid, aes(x = x, y = y, fill = category)) +
+    geom_tile(color = "black", size = 0.5) +
+    scale_fill_brewer(palette = "Set3") +
+    labs(title="Waffle Chart of Citations per Race")+
+    theme_void()
+}
+

### 2023-04-16 killer_plot1.Rmd

@@
 bar_run <- function(plot_race,plot_year){
-  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year), ]
-  new_df <- new_df[which(citations_htx_bar$subject_race == plot_race),]
+  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year &
+                                      citations_htx_bar$subject_race==plot_race), ]
   male_df <- new_df[which(new_df$subject_sex == "male"),]

### 2023-04-16 Final Presentation.Rmd
slide with shiny

---
title: "Final Presenation"
author: "Siddhi Narayan"
date: "2023-04-16"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

## Primary Dataset: Citations in HTX
Insert description

## Secondary Datasets
* Census Data
* GIS/Beats Data

## Census Data

## GIS/Beats Data

# Exploration

## Introduction to Exploration

## Understanding
* Plot - Latitude and Longitude of Stops

##
plot

## Distribution Over Time
* Plot - Line Graph of Citations by Race over Time
* Plot - Frequency of Citations vs. Hours in Day

##
plot

##
plot

## Distribution by Demographic
* Plot - Waffle Chart of Citation by Race 
* Plot - MPH over Speed Limit for Race/Gender

## 
plot

##
plot

## Distribution by Location
* Plot - Citation by Sex by District 
* Table - Citations Per Square Mile

## 
plot

## 
table

## Cumulative: Killer Plot
<iframe src="http://127.0.0.1:4571/" width="100%"></iframe>
# Analysis

## Introduction to Analysis

## Houston Racial Breakdown vs Citations Racial Breakdown
table

## Citation Type Racial Breakdown
table

## Failure to Establish Financial Responsibility
table

## Average Speeding Amounts
table

# Conclusion

## Conclusions
conclusions

### 2023-04-16 subscripts/speed_over_freq.R
updates for presentation

@@ -3,9 +3,9 @@ source("shiny_plot_setup.R")
 #frequency of different levels
 speed_over_freq <- function(){
 
-  mini_htx2 <- mini_htx[seq(1,666666,30),]
-  mini_htx2$district <- factor(mini_htx2$district)
-  mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
+  #mini_htx2 <- mini_htx[seq(1,666666,30),]
+  #mini_htx2$district <- factor(mini_htx2$district)
+  #mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
   mini_htx3 <- mini_htx2[, c("date", "diff")]
   mini_htx3 <- na.omit(mini_htx3)
   mini_htx3 <- mini_htx3[mini_htx3$diff != 693,]

### 2023-04-16 subscripts/violin_speed.R

@@ -1,8 +1,13 @@
 source("shiny_plot_setup.R")
 
+
+#mini_htx2$district <- factor(mini_htx$district)
+#mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
+
 violin_speed <- function(){
-  mini_htx4 <- na.omit(mini_htx2[,c("subject_race", "date", "diff", "subject_sex")])
+  mini_htx4 <- na.omit(mini_htx[,c("subject_race", "date", "diff", "subject_sex")])
   mini_htx4 <- mini_htx4[mini_htx4$diff != 693,]
+  mini_htx4 <- mini_htx4[mini_htx4$diff <= 105,]
   
   ggplot(mini_htx4, aes(fill=factor(subject_sex),subject_race, diff)) + 
     geom_violin() + 

### 2023-04-16 Final Presentation.Rmd
help

@@
 date: "2023-04-16"
-output: ioslides_presentation
+output: 
+  ioslides_presentation:
+    widescreen: true
+    css: style.css
 runtime: shiny
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, fig.width = 12, fig.height = 8, out.width = "100%")
 ```
@@
 table
+---
+## Cumulative: Killer Plot
+```{r, echo= FALSE, message= FALSE}
+source("shiny_plot_setup.R")
 
-## Cumulative: Killer Plot
-<iframe src="http://127.0.0.1:4571/" width="100%"></iframe>
+# Define UI for application that draws a histogram
+shinyUI(fluidPage(
+
+    # Application title
+    titlePanel("Citation Breakdown by District"),
+
+    # Sidebar with a slider input for number of bins
+    sidebarLayout(
+        sidebarPanel(
+          sliderInput("districts", "District:",
+                      min = 1, max = 24,
+                      value = 1, step = 1)
+        ),
+
+        # Show a plot of the generated distribution
+        mainPanel(
+            plotOutput("killerplot")
+        )
+    )
+))
+#<iframe src="http://127.0.0.1:4571/" width="100%" height="500px"></iframe>
+```
 # Analysis

### 2023-04-16 Presentation.Rmd

@@
 knitr::opts_chunk$set(echo = FALSE)
+source("shiny_plot_setup.R", echo=FALSE)
+#show_col_types = FALSE
 ```
@@
 
---- 
-plots
 - plot of citations by race over time  
@@
 
----
-plot
+## Pyramid 
+```{r, message = FALSE}
+source("subscripts/citation_pyramid.R", local = TRUE, print.eval = FALSE, echo= FALSE, verbose= FALSE)
+pyramid_cites()
+```
 
----
-
-```{r}
-#source("plots/plot.r", local = TRUE, print.eval = TRUE)
+## Violin
+```{r, message = FALSE}
+source("subscripts/violin_speed.R")
+violin_speed()
 ```
@@
 ---
+
+#Race Line Graph by Year: line_ry()
+#waffle plot: waffle()

### 2023-04-17 server.R
make killer fit

@@ -109,6 +109,6 @@ shinyServer(function(input, output) {
           popViewport()
         }
       }
-    }, width=800, height=800)
+    }, width=600, height=450)
 
 })

### 2023-04-17 Final Presentation.Rmd
killer on final presentation

--- 2023-04-16 Final Presentation.Rmd
@@
 
-## Cumulative: Killer Plot
-<iframe src="http://127.0.0.1:4571/" width="100%"></iframe>
+## Killer 
+<iframe src="http://127.0.0.1:5771" width="100%" height="500px"></iframe>
+
 # Analysis

### 2023-04-17 server.R
update push

@@ -38,7 +38,7 @@ shinyServer(function(input, output) {
       grid.lines(c(1.75/8, 0.9), c(1.5/8, 1.5/8))
       grid.lines(c(1.75/8, 1.75/8), c(1.5/8, 0.9))
       #grid.lines(c(0.1, 0.9), c(0.5, 0.5))
-      grid.text("Killer Plot",
+      grid.text("Citation Breakdown by Sex, Race, & Year", #Citation Breakdown by Sex for each Race & Year 
                 x = 1.75/8, y = 0.925, just = c("left", "bottom"),
                 gp = gpar(fontsize = 18)
       )

### 2023-04-17 subscripts/line_race_year.R

@@ -3,14 +3,14 @@
 
 source("shiny_plot_setup.R")
 
-race_date <- mini_htx[,c("date", "subject_race")] # race and date
-race_date <- na.omit(race_date) # omit na Values
-
 year_only <- function(fulldate){
   year <- substring(fulldate, 1, 4) # only keep year
 }
 
 line_ry <- function(){ 
+  race_date <- mini_htx[,c("date", "subject_race")] # race and date
+  race_date <- na.omit(race_date) # omit na Values
+  
   black <- race_date[race_date$subject_race=="black",]
   black$year <- unlist(lapply(black$date, year_only))
   plot(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), 

### 2023-04-17 Tables/Racial_Breakdown.R
tables asf

@@ -1,0 +1,19 @@
+source("shiny_plot_setup.R")
+# Question 1
+htx_stats <- read.csv("Datasets/Census Data Houston.csv")
+htx_filtered <- filter(htx_stats, `Fact` %in% c("White alone, percent","Black or African American alone, percent", "American Indian and Alaska Native alone, percent", "Asian alone, percent", "Native Hawaiian and Other Pacific Islander alone, percent","Two or More Races, percent"))
+htx_filtered <- data.frame(htx_filtered$Fact, htx_filtered$Houston.city..Texas)
+htx_filtered <- htx_filtered %>% rename(race = htx_filtered.Fact, population = htx_filtered.Houston.city..Texas)
+htx_filtered <- htx_filtered %>%
+  mutate(pop = as.numeric(sub("%", "", population))/100)  %>%
+  select(-population)
+
+aapi <- htx_filtered %>%filter(`race` %in% c("Asian alone, percent", "Native Hawaiian and Other Pacific Islander alone, percent"))%>% summarize(race = "asian/pacific islander", pop = sum(`pop`))
+htx_filtered <- htx_filtered %>% filter(!race %in% c("Asian alone, percent","Native Hawaiian and Other Pacific Islander alone, percent"))
+htx_filtered <- bind_rows(htx_filtered, aapi)
+htx_pop <- htx_filtered %>% mutate(race = race %>% recode("White alone, percent" = "white", "Black or African American alone, percent" = "black", "American Indian and Alaska Native alone, percent" = "american indian/alaska native", "Two or More Races, percent" = "other"))
+counts <- data.frame(na.omit(table(mini_htx$subject_race)))
+counts$Freq <- counts$Freq/sum(counts$Freq)
+counts <- counts %>% rename(race = Var1, citation_prop = Freq)
+htx <- inner_join(htx_pop, counts, by = "race")
+

### 2023-04-17 Tables/Racial_Citation.R

@@ -1,0 +1,32 @@
+source("shiny_plot_setup.R")
+speeding_race <- mini_htx %>% filter(grepl("SPEEDING", violation)) %>%
+  group_by(subject_race) %>%
+  summarise(n_individuals = n()) %>%
+  arrange(desc(n_individuals))
+
+# Invalid License Citations
+license_race <- mini_htx %>% filter(grepl("LICENSE", violation)) %>%
+  group_by(subject_race) %>%
+  summarise(n_individuals = n()) %>%
+  arrange(desc(n_individuals))
+
+
+# Failure to establish Financial Responsibility Citations
+financial_race <- mini_htx %>% filter(grepl("FINANCIAL RESPONSIBILITY", violation)) %>%
+  group_by(subject_race) %>%
+  summarise(n_individuals = n()) %>%
+  arrange(desc(n_individuals))
+
+# seat belt citations
+seat_belt_race <- mini_htx %>% filter(grepl("SEAT BELT", violation)) %>%
+  group_by(subject_race) %>%
+  summarise(n_individuals = n()) %>%
+  arrange(desc(n_individuals))
+
+
+# Running a stop light/red light
+stop_light_race <- mini_htx %>% 
+  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)) %>%
+  group_by(subject_race) %>%
+  summarise(n_individuals = n()) %>%
+  arrange(desc(n_individuals))

### 2023-04-17 Final Presentation.Rmd

@@
 title: "Final Presenation"
-author: "Siddhi Narayan"
-date: "2023-04-16"
+author: "Siddhi Narayan, Grace Wang, Virginia"
+date: "2023-04-17"
 output: 
@@
 
-## Distribution Over Time
+# Distribution Over Time
 * Plot - Line Graph of Citations by Race over Time
@@
 
-##
-plot
+## Citation Count by Race over the Years
+```{r Race/Year Line, message=FALSE}
+source("subscripts/line_race_year.R", local=TRUE, print.eval=TRUE)
+line_ry()
+```
 
-##
-plot
+## Citation Frequency for Time of Day
+```{r Time of Day Freq, message=FALSE}
+source("subscripts/line_timeofcites.R", local=TRUE, print.eval=TRUE)
+time_line()
+```
 
-## Distribution by Demographic
+# Distribution by Demographic
 * Plot - Waffle Chart of Citation by Race 
@@
 
-## 
-plot
+## Citation by Race
+* How close is this to the truth?
+```{r Waffle, message=FALSE}
+source("subscripts/waffle.R", local=TRUE, print.eval=TRUE)
+waffle()
+```
 
-##
-plot
+## MPH over Speed Limit for Race & Gender {.smaller}
+```{r Violin, message=FALSE}
+source("subscripts/violin_speed.R", local=TRUE, print.eval=TRUE)
+violin_speed()
+```
 
-## Distribution by Location
+# Distribution by Location
 * Plot - Citation by Sex by District 
@@
 
-## 
-plot
+## Citation by Sex by District
+```{r Pyramid, message = FALSE}
+source("subscripts/citation_pyramid.R", local = TRUE, print.eval = TRUE)
+pyramid_cites()
+```
 
-## 
-table
+## Citations Per Square Mile
+* table
 
-## Killer 
+## Killer Plot
+* Further breakdown Killer Plot by Police District
 <iframe src="http://127.0.0.1:5771" width="100%" height="500px"></iframe>

### 2023-04-17 Final Presentation.Rmd
update

@@
     widescreen: true
-    css: style.css
 runtime: shiny
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.width = 12, fig.height = 8, out.width = "100%")
+knitr::opts_chunk$set(echo = FALSE, fig.width = 10, fig.height = 6, out.width = "100%")
 ```
@@
 ## Citation by Race 
-* How close is this to the truth?
 ```{r Waffle, message=FALSE}
@@
 ## Killer Plot
-* Further breakdown Killer Plot by Police District
-<iframe src="http://127.0.0.1:5771" width="100%" height="500px"></iframe>
+<iframe src="http://127.0.0.1:5290" width="100%" height="500px"></iframe>
 

### 2023-04-17 draft4.Rmd
update push
update

@@
 points(sort(unique(other$year)), as.vector(table(as.factor(other$year))))
-  lines(sort(unique(other$year)), as.vector(table((as.factor(other$year)))), 
-        type = "l", col = "purple")
+lines(sort(unique(other$year)), as.vector(table((as.factor(other$year)))), type = "l", col = "purple")
 
-legend("topright", legend=c("black", "white", "aapi", "unknown", "other"), col=c("blue", "red", "green", "orange", "purple"), pch = 20, cex = 0.7, title = "Race")
+graphics::legend("topright", 
+       legend=c("black", "white", "aapi", "unknown", "other"), 
+       col=c("blue", "red", "green", "orange", "purple"), 
+       pch = 20, cex = 0.7, title = "Race")
 
@@
 plot(mini_htx2$lat, mini_htx2$lng, col = mini_htx2$district, panel.first = grid(8,8),  xlim = c(29.5, 30.15), ylim = c(-96, -95), main = "Latitude & Longitude of Stops by District", xlab = "Latitude", ylab = "Longitude", pch=1)
-legend("bottomright", legend=levels(mini_htx2$district), cex = 0.62, pch=1, col=unique(mini_htx2$district), title = "District")
+graphics::legend("bottomright", legend=levels(mini_htx2$district), cex = 0.62, pch=1, col=unique(mini_htx2$district), title = "District")
 ```

### 2023-04-18 Final Presentation.Rmd
update

--- 2023-04-17 Final Presentation.Rmd
@@
 ---
-title: "Final Presenation"
-author: "Siddhi Narayan"
-date: "2023-04-16"
+title: "Final Presentation"
+author: "Siddhi Narayan, Grace Wang, Virginia Baskin"
+date: "2023-04-17"
 output: 
   ioslides_presentation:
+    smaller: true
     widescreen: true
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.width = 12, fig.height = 8, out.width = "100%")
+knitr::opts_chunk$set(echo = FALSE, fig.width = 9, fig.height = 5, out.width = "100%")
 ```
@@
 
-## Census Data
+## Exploration
+- bullet 
+- point
 
-## GIS/Beats Data
+# Distribution Over Time
 
-# Exploration
+## Citation Count by Race over the Years {.smaller}
+```{r Race/Year Line, message=FALSE}
+source("subscripts/line_race_year.R", local=TRUE, print.eval=TRUE)
+line_ry()
+```
 
-## Introduction to Exploration
+## Citation Frequency for Time of Day {.smaller}
+```{r Time of Day Freq, message=FALSE}
+source("subscripts/line_timeofcites.R", local=TRUE, print.eval=TRUE)
+time_line()
+```
 
-## Understanding
-* Plot - Latitude and Longitude of Stops
+# Distribution by Demographic
 
-##
-plot
+## Citation by Race {.smaller}
+```{r Waffle, message=FALSE}
+source("subscripts/waffle.R", local=TRUE, print.eval=TRUE)
+waffle()
+```
 
-## Distribution Over Time
-* Plot - Line Graph of Citations by Race over Time
-* Plot - Frequency of Citations vs. Hours in Day
+## MPH over Speed Limit for Race & Gender {.smaller}
+```{r Violin, message=FALSE}
+source("subscripts/violin_speed.R", local=TRUE, print.eval=TRUE)
+violin_speed()
+```
 
-##
-plot
-
-##
-plot
-
-## Distribution by Demographic
-* Plot - Waffle Chart of Citation by Race 
-* Plot - MPH over Speed Limit for Race/Gender
-
-## 
-plot
-
-##
-plot
-
-## Distribution by Location
+# Distribution by Location
 * Plot - Citation by Sex by District 
@@
 
-## 
-plot
+## Citation by Sex by District {.smaller}
+```{r Pyramid, message = FALSE}
+source("subscripts/citation_pyramid.R", local = TRUE, print.eval = TRUE)
+pyramid_cites()
+```
 
-## 
-table
+## Citations Per Square Mile {.smaller}
+* table
+* table 
+* table
 
-## Killer 
-<iframe src="http://127.0.0.1:5771" width="100%" height="500px"></iframe>
+## Killer Plot
+<iframe src="http://127.0.0.1:5290" width="100%" height="500px"></iframe>
 
@@
 
-## Introduction to Analysis
-
-## Houston Racial Breakdown vs Citations Racial Breakdown
+## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
 table
 
-## Citation Type Racial Breakdown
+## Citation Type Racial Breakdown {.smaller}
 table
 
-## Failure to Establish Financial Responsibility
+## Failure to Establish Financial Responsibility {.smaller}
 table
 
-## Average Speeding Amounts
+## Average Speeding Amounts {.smaller}
 table

### 2023-04-18 Final Presentation.Rmd
slay

@@
 knitr::opts_chunk$set(echo = FALSE, fig.width = 9, fig.height = 5, out.width = "100%")
+source("shiny_plot_setup.R", echo=FALSE)
+source("Tables/Racial_Citation.R", echo = FALSE)
+source("Tables/Racial_Breakdown.R", echo = FALSE)
+source("Tables/FinResp.R", echo = FALSE)
+source("Tables/average_speeding_amt.R", echo = FALSE)
+source("Tables/csqm.R", echo = FALSE)
 ```
@@
 ## Citations Per Square Mile {.smaller}
-* table
-* table 
-* table
+Top 5 Beats ranked by citations/square mile:
+```{r csqm_top, echo = FALSE, message = FALSE}
+knitr::kable(csqm_top)
+```
+##
+Bottom 5 Beats ranked by citations/square mile
+```{r csqm_bottom, echo = FALSE, message = FALSE}
+knitr::kable(csqm_bottom)
+```
+##
+Statistics on number of citations/square mile:
+
+```{r csqm_summry, echo = FALSE, message = FALSE}
+knitr::kable(cqm_summary)
+```
 
@@
 ## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
-table
+```{r racial_citation_breakdown, echo = FALSE, message = FALSE}
+knitr::kable(htx)
+```
 
 ## Citation Type Racial Breakdown {.smaller}
-table
+Speeding Citations by Race:
+```{r speeding_race, echo = FALSE, message = FALSE}
+knitr::kable(speeding_race)
+```
+##
+Invalid License Citations by Race:
+```{r license_race, echo = FALSE, message = FALSE}
+knitr::kable(license_race)
+```
+
+##
+Failure to Establish Financial Responsibility Citations by Race:
+```{r financial_race, echo = FALSE, message = FALSE}
+knitr::kable(financial_race)
+```
+
+##
+Seat Belt Citations by Race:
+```{r seat_belt_race, echo = FALSE, message = FALSE}
+
+knitr::kable(seat_belt_race)
+```
+
+##
+Running a Red Light/Stop Sign Citations by Race:
+```{r stop_light_race, message = FALSE}
+knitr::kable(stop_light_race)
+```
 
 ## Failure to Establish Financial Responsibility {.smaller}
-table
+```{r fin_resp, echo = FALSE, message = FALSE}
+knitr::kable(all_perc)
+```
 
 ## Average Speeding Amounts {.smaller}
-table
+All People (baseline):
+```{r all_speed, echo = FALSE, message = FALSE}
+knitr::kable(all_speed)
+```
+##
+By Race:
+```{r race_speed, echo = FALSE, message = FALSE}
+knitr::kable(race_speed)
+```
+##
+By Gender:
+```{r gender_speed, echo = FALSE, message = FALSE}
+knitr::kable(gender_speed)
+```
+##
+```{r race_gender_speed, echo = FALSE, message = FALSE}
+library(kableExtra)
+kable(race_gender_speed) %>% kable
 

### 2023-04-18 draft4.Rmd
ignore

--- 2023-04-16 draft4.Rmd
@@
 mini_htx <- read_csv("Datasets/mini_htx.csv")
-full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
-census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
-mini_htx <- full_htx[seq(1,2000000,3),]
+# full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
+# census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
+# mini_htx <- full_htx[seq(1,2000000,3),]
 mini_htx <- mini_htx %>% rename("Beats" = "beat")

### 2023-04-18 Final Presentation.Rmd
still doing this

@@
 # Distribution by Location
-* Plot - Citation by Sex by District 
-* Table - Citations Per Square Mile
 
@@
 
-## Citations Per Square Mile {.smaller}
-Top 5 Beats ranked by citations/square mile:
+## Citations Per Square Mile | Top 5 Beats ranked by citations/square mile:
 ```{r csqm_top, echo = FALSE, message = FALSE}
@@
 ```
-##
-Bottom 5 Beats ranked by citations/square mile
+
+## Citations Per Square Mile | Bottom 5 Beats ranked by citations/square mile
 ```{r csqm_bottom, echo = FALSE, message = FALSE}
@@
 ```
-##
-Statistics on number of citations/square mile:
 
+## Citations Per Square Mile | Statistics on citations per square mile:
 ```{r csqm_summry, echo = FALSE, message = FALSE}
@@
 ## Killer Plot
-<iframe src="http://127.0.0.1:5290" width="100%" height="500px"></iframe>
+<iframe src="http://127.0.0.1:5501" width="100%" height="500px"></iframe>
 
@@
 library(kableExtra)
-kable(race_gender_speed) %>% kable
-
+kable(race_gender_speed) %>% kable_styling(font_size=8)
+``` 
 # Conclusion

### 2023-04-18 Tables/average_speeding_amt.R
not raw

@@ -7,7 +7,7 @@ all_speed <- mutate(mini_htx, diff = speed - posted_speed) %>%
             median = median(diff, na.rm = TRUE), 
             sd = sd(diff, na.rm = TRUE))
 race_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
-  filter(posted_speed >= 0) %>% group_by(raw_race) %>%
+  filter(posted_speed >= 0) %>% group_by(subject_race) %>%
   summarise(total = n(),  ## Function to use internally of summarise only
             mean = mean(diff, na.rm = TRUE),
             median = median(diff, na.rm = TRUE),
@@ -19,7 +19,7 @@ gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>%
             median = median(diff, na.rm = TRUE),
             sd = sd(diff, na.rm = TRUE)) 
 race_gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
-  filter(posted_speed >= 0) %>% group_by(raw_race, subject_sex) %>%
+  filter(posted_speed >= 0) %>% group_by(subject_race, subject_sex) %>%
   summarise(total = n(),
             mean = mean(diff, na.rm = TRUE),
             median = median(diff, na.rm = TRUE),

### 2023-04-18 Tables/FinResp.R
tables header changed

@@ -45,4 +45,4 @@ nas_fin <- nas %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBIL
 nas_perc <- nas_fin[1]/nas_total
 
 all_perc <- data.frame(c(blk_perc, white_perc, aapi_perc, unknown_perc, nas_perc))
-colnames(all_perc) <- c("black", "white", "aapi", "unknown","N/A")
+colnames(all_perc) <- c("Black", "White", "AAPI", "Unknown","Unrecorded")

### 2023-04-18 Tables/Racial_Citation.R

@@ -2,26 +2,30 @@ source("shiny_plot_setup.R")
 speeding_race <- mini_htx %>% filter(grepl("SPEEDING", violation)) %>%
   group_by(subject_race) %>%
   summarise(n_individuals = n()) %>%
-  arrange(desc(n_individuals))
+  arrange(desc(n_individuals)) %>%
+  rename("Subject Race"="subject_race", "Speeding"="n_individuals")
 
 # Invalid License Citations
 license_race <- mini_htx %>% filter(grepl("LICENSE", violation)) %>%
   group_by(subject_race) %>%
   summarise(n_individuals = n()) %>%
-  arrange(desc(n_individuals))
+  arrange(desc(n_individuals))%>%
+  rename("Subject Race"="subject_race", "Invalid License"="n_individuals")
 
 
 # Failure to establish Financial Responsibility Citations
 financial_race <- mini_htx %>% filter(grepl("FINANCIAL RESPONSIBILITY", violation)) %>%
   group_by(subject_race) %>%
   summarise(n_individuals = n()) %>%
-  arrange(desc(n_individuals))
+  arrange(desc(n_individuals))%>%
+  rename("Subject Race"="subject_race", "Financial Responsibility"="n_individuals")
 
 # seat belt citations
 seat_belt_race <- mini_htx %>% filter(grepl("SEAT BELT", violation)) %>%
   group_by(subject_race) %>%
   summarise(n_individuals = n()) %>%
-  arrange(desc(n_individuals))
+  arrange(desc(n_individuals))%>%
+  rename("Subject Race"="subject_race", "Seat Belt"="n_individuals")
 
 
 # Running a stop light/red light
@@ -29,9 +33,11 @@ stop_light_race <- mini_htx %>%
   filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)) %>%
   group_by(subject_race) %>%
   summarise(n_individuals = n()) %>%
-  arrange(desc(n_individuals))
+  arrange(desc(n_individuals))%>%
+  rename("Subject Race"="subject_race", "Stop Sign/Red Light"="n_individuals")
 
-citation_race <- inner_join(speeding_race, license_race, by = "subject_race")
-citation_race <- inner_join(citation_race, financial_race, by = "subject_race")
-citation_race <- inner_join(citation_race, seat_belt_race, by = "subject_race")
-citation_race <- inner_join(citation_race, stop_light_race, by = "subject_race")
[…]

### 2023-04-18 Final Presentation.Rmd

@@
 ## Primary Dataset: Citations in HTX
-Insert description
+- Citation Data in Houston, obtained from Stanford Open Policing Project
+- Consists of collection of over 2,000,000 stops over various variables
 
@@
 ## Exploration
-- bullet 
-- point
+- Understand and contextualize primary dataset
+- Investigate trends that may exhibit potential bias
+- Examined distribution of citations over time, distribution by demographic and distribution by location
 

### 2023-04-18 Final Presentation.Rmd
final presentation tidied up

--- 2023-04-18 Final Presentation.Rmd
@@
 ## Primary Dataset: Citations in HTX
-Insert description
+- Citation Data in Houston, obtained from Stanford Open Policing Project
+- Consists of collection of over 2,000,000 stops over various variables
+- Has 13 key fields of interest concerning the stop 
+  - Race,
+  - Sex, 
+  - Police District, 
+  - Citation Type, 
+  - Location (lat/long),
+  - Car Make and Model
+  - and so on...  
+  - Importantly, all but 3 field are categorical (Speed, Latitude, & Longitude being the only exception)
 
@@
 * Census Data
+  - Census Data for the City of Houston from 2021, accessed through US Census Bureau Website    
+  - Provides population estimates for different races and ethnicities in Houston    
+  - We primarily utilized population percentage estimates of race
+
+  <br>
 * GIS/Beats Data
+  - Contained GIS (location) data for the City of Houston's Police Beats    
+  - Police District and Area, along with other specific geographic features for each of the 118 Police Beats
 
 ## Exploration
-- bullet 
-- point
 
-# Distribution Over Time
+- Understand and contextualize primary dataset    
+- Investigate trends that may exhibit potential bias     
+- Examined distribution of citations over time, distribution by demographic and distribution by location
+
+# Distributions Over Time
 
@@
 
-# Distribution by Demographic
+# Distributions by Demographic
 
-## Citation by Race {.smaller}
+## Citation Count by Race {.smaller}
 ```{r Waffle, message=FALSE}
@@
 waffle()
+```
+
+## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
+```{r racial_citation_breakdown, echo = FALSE, message = FALSE}
+knitr::kable(htx)
 ```
@@
 
-# Distribution by Location
+# Distributions by Location
 
@@
 
+# Killer Plot!
+
 ## Killer Plot
@@
 
-# Analysis
-
-## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
-```{r racial_citation_breakdown, echo = FALSE, message = FALSE}
-knitr::kable(htx)
-```
+# Citation Specific Analysis
 
@@
 
+## CHANGE THE NAMES
+```{r mega table, message = FALSE}
+knitr::kable(citation_race)
+```
+
 ## Failure to Establish Financial Responsibility {.smaller}
@@
 ```
+
+## Financial Responsibility Hypothesis Testing {.smaller}
+
+We preformed a hypothesis test at $\alpha = 0.05$ (95% confidence) in order to identify which races have a significantly larger proportion of citations issues solely for "failure to establish financial responsibility" than other races.
+  <br>
+
+$H_0: p_1 = p_2$
+
+$H_1: p_1 > p_2$
+
+- Unrecorded Race (N/A) was greater than all other races 
+  - with $pval < 0.00001$
+
+- Black individuals had a higher proportion than white, unknown, and AAPI 
+  - with $pval < 0.0005$
 
@@
 kable(race_gender_speed) %>% kable_styling(font_size=8)
+detach(package:kableExtra, unload=TRUE)
 ```
+
 # Conclusion
@@
 ## Conclusions
-conclusions
+- We found preliminary evidences of bias:    
+  - In the differences of proportions by race for failure to show financial responsibility as the only citation     
+  - Black women were given speeding citations at roughly the same rate as black men, but the rate of speeding citations for women of all other races is consistently half that of their male counterparts
+
+  <br>    
+- Limitations to our analysis:    
+  - More information needed to confirm any of these findings (so much missing information)    
+    - Pattern to the missingness?     
+    - Severity of bias could change     
+  - We cannot exactly quantify individual bias (as everything is on an aggregate)
+
+# Questions?

### 2023-04-18 Tables/average_speeding_amt.R
table amendments

@@ -24,3 +24,6 @@ race_gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>%
             mean = mean(diff, na.rm = TRUE),
             median = median(diff, na.rm = TRUE),
             sd = sd(diff, na.rm = TRUE)) 
+race_gender_speed <- race_gender_speed %>%
+  filter_at("subject_sex", all_vars(!is.na(.))) %>%
+  rename("Subject Sex"="subject_sex", "Total"="total", "Mean"="mean", "Median"="median", "SD"="sd")

### 2023-04-18 Tables/csqm.R

@@ -20,4 +20,4 @@ cqm_summary[,3] <- csqm %>% summarise(IQR(Citations_per_sqm))
 
 csqm_top <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_head(n=5)
 
-csqm_bottom <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_tail(n=5)
+csqm_bottom <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_tail(n=5) %>% rename("Citations Per Sq. mi."="Citations_per_sqm")

### 2023-04-18 Tables/average_speeding_amt.R
slay

@@ -11,13 +11,15 @@ race_speed <- mutate(mini_htx, diff = speed - posted_speed) %>%
   summarise(total = n(),  ## Function to use internally of summarise only
             mean = mean(diff, na.rm = TRUE),
             median = median(diff, na.rm = TRUE),
-            sd = sd(diff, na.rm = TRUE)) 
+            sd = sd(diff, na.rm = TRUE))%>%
+  rename("Subject Race"="subject_race") 
 gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
   filter(posted_speed >= 0) %>% group_by(subject_sex) %>%
   summarise(total = n(),  
             mean = mean(diff, na.rm = TRUE),
             median = median(diff, na.rm = TRUE),
-            sd = sd(diff, na.rm = TRUE)) 
+            sd = sd(diff, na.rm = TRUE)) %>%
+  rename("Subject Sex"="subject_sex")
 race_gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
   filter(posted_speed >= 0) %>% group_by(subject_race, subject_sex) %>%
   summarise(total = n(),
@@ -26,4 +28,5 @@ race_gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>%
             sd = sd(diff, na.rm = TRUE)) 
 race_gender_speed <- race_gender_speed %>%
   filter_at("subject_sex", all_vars(!is.na(.))) %>%
-  rename("Subject Sex"="subject_sex", "Total"="total", "Mean"="mean", "Median"="median", "SD"="sd")
+  rename("Subject Sex"="subject_sex", "Subject Race" = "subject_race", "Total"="total", "Mean"="mean", "Median"="median", "SD"="sd")
+race_gender_speed

### 2023-04-18 Tables/csqm.R

@@ -18,6 +18,6 @@ cqm_summary[,1] <- csqm %>% summarise(median(Citations_per_sqm))
 cqm_summary[,2] <- csqm %>% summarise(sd(Citations_per_sqm))
 cqm_summary[,3] <- csqm %>% summarise(IQR(Citations_per_sqm))
 
-csqm_top <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_head(n=5)
+csqm_top <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_head(n=5) %>% rename("Citations Per Sq. mi."="Citations_per_sqm")
 
 csqm_bottom <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_tail(n=5) %>% rename("Citations Per Sq. mi."="Citations_per_sqm")

### 2023-04-19 FinalReport.Rmd
Compile Final Report

---
title: "Investigating for Bias in Houston, Texas Police Citations"
author: "Grace Wang, Virginia Baskin, Siddhi Narayan"
date: "2023-04-18"
output: pdf_document
fontsize: 10pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load packages, include=FALSE}
library(RSQLite)
library(readr)
library(ggplot2)
library(ggExtra)
library(ggthemes)
library(dplyr)
library(knitr)

mini_htx <- read_csv("Datasets/mini_htx.csv")
# full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
# census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
# mini_htx <- full_htx[seq(1,2000000,3),]
mini_htx <- mini_htx %>% rename("Beats" = "beat")
beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
# htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")
#dbWriteTable(con, "census", census_csv, overwrite = TRUE)

```

# Introduction

## Problem and Importance
For our project, we wanted to explore vehicular police stop data from Houston, Texas to investigate and identify any potential biases against particular populations. This problem is not trivial, as it could contribute towards revealing patterns of discrimination or disproportionate targeting of certain groups based on race or gender. Such biases can have serious consequences, such as perpetuating systemic inequality and eroding trust in law enforcement. By analyzing data on police stops and citations, we can illuminate any incongruencies we find to hopefully influence targeted interventions, such as training programs, policy changes, or even community policing initiatives. Overall, investigating data on police stops is an essential step towards creating a fair and just society that ensures equal treatment and protection under the law for all individuals, regardless of their demographic.

## Primary Dataset: Police Citations in Houston, Texas
The primary dataset that we used was obtained from the Stanford Open Policing Project, which is a project based in Stanford University that works on collecting and standardizing data on vehicle and pedestrian stops from law enforcement departments across the country. For this project, we looked at the data based in Houston. 
This dataset consists of a collection of 2,045,972 stops, providing us with an extensive amount of data to work with. The information collected in this dataset typically includes the date and time of the stop, the address of the stop, the latitude and longitude coordinates, and the geocode of the stop. We also have information on the police beat and police district where the stop was performed.
Additionally, if the stop was a vehicular stop, the dataset includes data on the speed of the vehicle and the speed limit of the area, which can be helpful in understanding the extent to which traffic violations are being targeted by law enforcement officers. We also have information on the color, make, and model of the vehicle. Finally, the dataset includes information on whether a citation was issued and what the outcome of the stop was. 
One thing to note is that most of these variables (all but three) were categorical, so we had to find unique and innovative ways to analyze and contextualize them.

## Secondary Datasets: Census and GIS Data
### Census Data
One of the supplementary datasets that our team used was the Census Data of the City of Houston from 2021, which we accessed through the United States Census Bureau website. This dataset gave population estimates of Houston, population percentage estimates of Age and Sex, population percentage estimates of Race and Hispanic Origin, estimates of Population Characteristics, and more. For the purpose of this report, we focused on the population percentage estimates of Race, which gave percentage estimates of White alone, Black or African American alone, American Indian and Alaska Native alone, Asian alone, Native Hawaiian and Other Pacific Islander alone and Two or More Races.

### GIS/Beats Data
Another supplementary dataset our team utilized was GIS Data for the City of Houston Police Beats. This dataset contains 118 rows (one for each police beat in Houston) and has columns for police district and area (in square miles) among other geographical characteristics. We paired this dataset with our primary dataset in order to geographically contextualize our findings.

# Exploration
The purpose of our exploration was to understand and contextualize the large primary dataset and investigate any trends that might exhibit potential for bias. We did this by examining a few different factors: distribution  of citations over time, by demographic, and by location. 
For distribution of citations over time, we aimed to see if there were any discrepancies between citations for each race over time, and generally see how citations differed from 2014-2019. We also looked at the frequency of citations per hour to see if there were any particular patterns, as well as the miles per hour over speed limit per year. 
For distribution by demographic, we examined the data in several ways: a waffle chart of citation by race, a table comparing citation proportions for each race to their population proportion in Houston, and a plot of MPH over Speed Limit for Race/Gender.
For distribution by location, we examined this through a population plot of citations by sex and district, as well as a table that displays the number of citations per square mile.
Finally, we combined our investigation of time, demographic, and location in our killer plot, which gave us insight into the distribution of citations by race, gender, and district over time.

## Distribution Over Time
```{r Race Line Graph by Year, echo=FALSE}
# Visualize totals for each race by year
# Grace

race_date <- mini_htx[,c("date", "subject_race")] # race and date
race_date <- na.omit(race_date) # omit na Values

year_only <- function(fulldate){
  year <- substring(fulldate, 1, 4) # only keep year
}

black <- race_date[race_date$subject_race=="black",]
black$year <- unlist(lapply(black$date, year_only))
plot(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), ylim = c(0, 60000), main = "Citations by Race for Each Year", xlab="Year", ylab = "Count") #200000 when all
lines(sort(unique(black$year)), as.vector(table((as.factor(black$year)))), type = "l", col = "blue")

white <- race_date[race_date$subject_race=="white",]
white$year <- unlist(lapply(white$date, year_only))
points(sort(unique(white$year)), as.vector(table(as.factor(white$year))))
lines(sort(unique(white$year)), as.vector(table((as.factor(white$year)))), type = "l", col = "red")

aapi <- race_date[race_date$subject_race=="asian/pacific islander",]
aapi$year <- unlist(lapply(aapi$date, year_only))
points(sort(unique(aapi$year)), as.vector(table(as.factor(aapi$year))))
lines(sort(unique(aapi$year)), as.vector(table((as.factor(aapi$year)))), type = "l", col = "green")

unknown <- race_date[race_date$subject_race=="unknown",]
unknown$year <- unlist(lapply(unknown$date, year_only))
points(sort(unique(unknown$year)), as.vector(table(as.factor(unknown$year))))
lines(sort(unique(unknown$year)), as.vector(table((as.factor(unknown$year)))), type = "l", col = "orange")

other <- race_date[race_date$subject_race=="other",]
other$year <- unlist(lapply(other$date, year_only))
points(sort(unique(other$year)), as.vector(table(as.factor(other$year))))
lines(sort(unique(other$year)), as.vector(table((as.factor(other$year)))), type = "l", col = "purple")

graphics::legend("topright", 
       legend=c("black", "white", "aapi", "unknown", "other"), 
       col=c("blue", "red", "green", "orange", "purple"), 
       pch = 20, cex = 0.7, title = "Race")

```
First, we evaluated the consistency of the distribution of citations between races over the years 2014-2020. We found that the racial distribution of citations remained fairly consistent: from 2014-2019, the greatest number of individuals issued citations were white, but in 2020, black individuals narrowly surpassed white individuals as the racial group with the most citations of any racial group. In each year, individuals of unknown race and asian/pacific islanders also received significantly fewer citations than white and black individuals. In 2019 and 2020, a decrease in citations issued to all races can be observed, likely due to the onset of the COVID-19 Pandemic.
From this visualization, we did not conclude that there were any significant changes in the relative distribution of citations issued to each racial group over time.

```{r Line graph of number of citations vs. Hour in Day, echo=FALSE}
# Observe citations over time and any time-associate patterns
time <- as.POSIXct(mini_htx$time, format = "%H:%M:%S")
hour <- as.numeric(format(time, "%H"))
count <- data.frame(table(na.omit(hour)))
plot(count$Var1, count$Freq, type = "n", xlab = "Hour", ylab = "Frequency of Citations", main = "Frequency of Citations vs Hour in Day")
lines(count$Var1, count$Freq, type = "l", lwd = 2)
```

Based on the analysis of the frequency of citations over the hours of the day, it appears that there are two distinct peaks, one around 8 am and another around 3 pm. There are several possible explanations for these peaks.
One possible explanation for the peak around 8 am is that it is due to high travel volume during rush hour. Many people are commuting to work at this time, which may increase the likelihood of traffic violations and, consequently, citations. This is further supported by the fact that there is a lull in citations during the super late and super early hours when there is typically less travel.
Another possible explanation for the peak around 3 pm is that it is related to school schedules. Children in K-12 typically get out of school around 3 pm, which may cause a spike in traffic volume as parents rush to pick up their children and return home before rush hour. This increased traffic volume could lead to a higher number of traffic violations and citations.

## Distribution by Demographic

```{r Waffle Plot of races, echo=FALSE}
df <- data.frame(table(na.omit(mini_htx$subject_race)))
total_freq <- sum(df$Freq)
df$prop <- df$Freq/total_freq
num_squares <- round(df$prop*100)

df_grid <- data.frame(
 x = rep(1:10, each = 10),
 y = rep(1:10, times = 10)
)
num_squares <- c(3, 36, 1, 3, 57) #manually add one more square because the rounding resulted in only 99 
#num_squares <- round(df$prop*100)

df_grid$category <- rep(df$Var1, num_squares)
ggplot(df_grid, aes(x = x, y = y, fill = category)) +
 geom_tile(color = "black", size = 0.5) +
 scale_fill_brewer(palette = "Set3") +
 labs(title="Waffle Chart of Citations per Race")+
 theme_void()
```
This representation of the percentage of citations by race across all years 2014-2020 provides a visual and geometric way to understand the distribution of citations across different racial groups. The graph shows that white individuals have the highest percentage of citations, followed by black individuals, and then Asian/Pacific Islanders and unknown races.
Following this plot, our team determined that we would like to compare the distribution of citations by race to the demographics of Houston to see if any racial groups receive a disproportionate amount of citations as compared to their representation within the city of Houston.

```{r Race Citation Proportion Table, echo=FALSE}
htx_stats <- read.csv("Datasets/Census Data Houston.csv")
htx_filtered <- filter(htx_stats, `Fact` %in% c("White alone, percent","Black or African American alone, percent", "American Indian and Alaska Native alone, percent", "Asian alone, percent", "Native Hawaiian and Other Pacific Islander alone, percent","Two or More Races, percent"))
htx_filtered <- data.frame(htx_filtered$Fact, htx_filtered$Houston.city..Texas)
htx_filtered <- htx_filtered %>% rename(race = htx_filtered.Fact, population = htx_filtered.Houston.city..Texas)
htx_filtered <- htx_filtered %>%
  mutate(pop = as.numeric(sub("%", "", population))/100)  %>%
  select(-population)

aapi <- htx_filtered %>%filter(`race` %in% c("Asian alone, percent", "Native Hawaiian and Other Pacific Islander alone, percent"))%>% summarize(race = "asian/pacific islander", pop = sum(`pop`))
htx_filtered <- htx_filtered %>% filter(!race %in% c("Asian alone, percent","Native Hawaiian and Other Pacific Islander alone, percent"))
htx_filtered <- bind_rows(htx_filtered, aapi)
htx_pop <- htx_filtered %>% mutate(race = race %>% recode("White alone, percent" = "white", "Black or African American alone, percent" = "black", "American Indian and Alaska Native alone, percent" = "american indian/alaska native", "Two or More Races, percent" = "other"))
counts <- data.frame(na.omit(table(mini_htx$subject_race)))
counts$Freq <- counts$Freq/sum(counts$Freq)
counts <- counts %>% rename(race = Var1, citation_prop = Freq)
htx <- inner_join(htx_pop, counts, by = "race")
htx <- htx %>%
  rename("Race"="race", "Population Proportion"="pop", "Citation Proportion"="citation_prop")
kable(htx)
```
This table shows that the citation proportion for White people is higher than their population proportion in Houston, indicating that they may be disproportionately cited. Similarly, the citation proportion for Black people is also higher than their population proportion, suggesting that they too may be disproportionately cited.
On the other hand, the citation proportion for Asian/Pacific Islander people is lower than their population proportion, indicating that they may be underrepresented in the citations. However, it's essential to keep in mind that the exact reasons for these disparities are not clear, and there may be several factors contributing to them, such as socioeconomic factors or geographical location.
It's also worth noting that the "other" population proportion is significantly higher than the citation proportion. Still, this may be due to the fact that mixed-race individuals are classified as "other," while for citation purposes, people giving citations may mark them as a single race.

```{r mph speeding violin plot, echo=FALSE}
mini_htx2 <- mini_htx
mini_htx2$diff <- abs(mini_htx2$speed - mini_htx2$posted_speed)
mini_htx4 <- na.omit(mini_htx2[,c("subject_race", "date", "diff", "subject_sex")])
  mini_htx4 <- mini_htx4[mini_htx4$diff != 693,]
  mini_htx4 <- mini_htx4[mini_htx4$diff <= 105,]

  ggplot(mini_htx4, aes(fill=factor(subject_sex),subject_race, diff)) + 
    geom_violin() + 
    labs(title="MPH over Speed Limit for Citation", x="Subject Race",y="MPH Speeding")+
    guides(fill=guide_legend(title='Subject Sex'))

```
Moving forward, we chose to visualize MPH over Speed Limit for Citation by subject race and sex in order to identify any discrepancies between speeding amounts that warrant a ticket for different races and genders.
We observed that the distribution of speeding amounts for individuals of white, black, and unknown race were similar, with the greatest peak around 10 mph, followed by another at 15 mph. Meanwhile, asian/pacific islanders demonstrated a more uniform distribution of speeding amounts from 10-20 mph that tapered off to a maximum of around 30 mph and a minimum of 5 mph. Lastly, individuals of “other” race received citations for speeding in amounts relatively uniformly distributed from 5-30 mph for men and 10-30 mph for women. Density of speeding amounts for each sex within each race were also very similar, although males of every race besides “other” reached a greater maximum speeding amount than females. This is likely due to differences in driving behaviors between males and females (with males typically having  risk tolerance than females), and not evidence of any biases.

## Distribution by Location
```{r Pyramid Plot, echo=FALSE, warning=FALSE}
#x axis citation number, y axis is district, color is gender

library(ggthemes)
options(scipen = 999)  # turns of scientific notations like 1e+40

test <- table(mini_htx$subject_sex, mini_htx$district)
new_df <- data.frame(matrix(ncol = 3, nrow = 0))
for (row in rownames(test)){
  for (col in colnames(test)){
    #row in new dataframe
    #(sex, district, value)
    new_df <- rbind(new_df, c(row, col, test[row, col]))
  }
}
colnames(new_df) <- c("sex", "district", "count")
new_df$count <- as.numeric(new_df$count)
new_df$district <- as.factor(new_df$district)

mens <- new_df[new_df$sex == "male",]
mens$count <- (mens$count * -1)
new_df <- rbind(new_df[new_df$sex == "female",], mens)
new_df$district <- ordered(new_df$district, seq(1,24))

brks <- seq(-50000, 30000, 10000)
lbls = paste0(as.character(c(seq(500, 0, -100), seq(100, 300, 100))), "k")
#lbls = paste0(as.character(c(seq(15, 0, -5), seq(5, 15, 5))), "m")

# Plot
library(ggfortify)
ggplot(new_df, aes(x = district, y = count, fill = sex)) +   # Fill column
                              geom_bar(stat = "identity", width = .7) +   # draw the bars
                              scale_y_continuous(breaks = brks, labels = lbls) + 
                              coord_flip() +  # Flip axes
                              labs(title="Citation by Sex by District - Pyramid", y = "Citation Count", x = "District") +
                              theme_tufte() +  # Tufte theme from ggfortify
                              theme(plot.title = element_text(hjust = .5), 
                                    axis.ticks = element_blank()) +   # Centre plot title
                              scale_fill_brewer(palette = "Dark2")  # Color palette
options(warn=-1)

```
To begin our investigation by location, we made a pyramidal graph of citation count by sex for each district. The x-axis is the amount of citations in thousands, and the y-axis is the police district in Houston. This plot reveals interesting aspects of our dataset regarding the proportion of citations by gender. There are generally more citations for men than women for each district. This could be because men are more likely to commit infractions due to higher risk tolerance, or it could be that the rates of infraction are the same for the genders and women are less likely to be given a citation. We think it is most likely the former influencing the imbalance. While the core reason for the discrepancy is unknown, the unequal distribution is clear to see. 
	An interesting feature is that Districts 21 and 23 have significantly less citations overall, and that is because they are Hobby airport and IAH airport, respectively. Since it is not residential and technically doesn’t have a population, it is only logical that there are less citations for those districts. 
	District 2 has the most citations overall, which is one of the most densely populated districts, as it contains Montrose, Greater Heights, West University, and other densely populated neighborhoods within the inner loop of Houston. The inner loop (roughly Districts 1, 2, 9, and 10) have ½ million people in ~90 square miles, so it makes sense to see an increase of citations in those districts.

```{r Citations per sqm table, echo=FALSE, message = FALSE}
beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
citations_in_beat <- mini_htx %>% inner_join(beats_data, by="Beats") %>%
  group_by(Beats) %>%
  summarise(n_individuals = n())

citations_in_beat <- as.data.frame(citations_in_beat)

cit_beat_area <- citations_in_beat %>% 
  inner_join(beats_data[,c("Beats", "Area_sq_mi")], by="Beats")

cit_beat_area$Citations_per_sqm <- cit_beat_area$n_individuals / cit_beat_area$Area_sq_mi
csqm <- cit_beat_area[,c("Beats","Citations_per_sqm")]

cqm_summary <- data.frame(matrix(ncol=3,nrow=1))
colnames(cqm_summary) <- c("Median","Standard Deviation", "IQR")
cqm_summary[,1] <- csqm %>% summarise(median(Citations_per_sqm))
cqm_summary[,2] <- csqm %>% summarise(sd(Citations_per_sqm))
cqm_summary[,3] <- csqm %>% summarise(IQR(Citations_per_sqm))

csqm_top <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_head(n=5) %>% rename("Citations Per Sq. mi."="Citations_per_sqm")

csqm_bottom <- csqm %>% arrange(desc(Citations_per_sqm)) %>% slice_tail(n=5) %>% rename("Citations Per Sq. mi."="Citations_per_sqm")

attr(cqm_summary, "title") <- "Summary of Citations per Square Mile"
kable(cqm_summary)
kable(csqm_top)
kable(csqm_bottom)

```

Next, our team aimed to understand the geographic distribution of citations by calculating the number of citations per square mile within each police beat. We hoped that this investigation would help us identify any beats with an abnormal quantity of citations so that we could narrow down the scope of our investigation for bias. As a part of our investigation, we found the median, standard deviation, and interquartile range of the number of citations per square mile for all police beats and the 5 beats with the greatest and least number of citations per square mile.
Upon receiving the list of 5 police beats with the greatest number of citations, we cross-referenced a map of Houston’s police beats to view the location and characteristics of these beats. We found that all five of these beats are adjacent, small, and located in the metropolitan heart of Houston. Additionally, we noted that both beats 1A10 and 2A40 are very small in terms of square mileage, but contain two and one police stations, respectively; proximity to police stations could therefore explain the extremely high citations/square mile values for these beats.
Upon receiving the list of 5 police beats with the lowest number of citations, we once again cross-referenced a map of Houston’s police beats to understand the characteristics of these beats. Our team found that these five beats were located in Houston's suburbs and are all relatively large in terms of square mileage, which explains their low citations/square mile values.

## Killer Plot
```{r, echo=FALSE, results='hide', message=FALSE}
library(stringr)
library(grid)
library(readr)
# subset the data to only contain columns of interest
citations_htx <- mini_htx[,c("subject_race", "subject_sex", "date", "violation")]

# subset the data to only contain rows of violation interest
citations_htx <- citations_htx %>% 
  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)|
           grepl("SEAT BELT", violation)|grepl("FINANCIAL RESPONSIBILITY", violation)|
           grepl("LICENSE", violation)|grepl("SPEEDING", violation)) %>%
  mutate(date = format(date, format = "%Y"))

#create the circles dataframe which counts individuals in each race, year pairing
citations_htx_circle <- citations_htx %>%
  group_by(subject_race, date) %>%
  summarise(n_individuals = n())

# create the bars dataframe which counts the number of individuals in each race, year, violation, pairing

# create a dataframe with all possible combinations of race, date, sex, and violation
citations_htx_bar <- citations_htx %>%
  mutate(red_stop= if_else(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation), 1, 0)) %>%
  mutate(seat_belt= if_else(grepl("SEAT BELT", violation), 1, 0)) %>%
  mutate(fin_resp= if_else(grepl("FINANCIAL RESPONSIBILITY", violation), 1, 0))%>%
  mutate(license= if_else(grepl("LICENSE", violation), 1, 0))%>%
  mutate(speeding= if_else(grepl("SPEEDING", violation), 1, 0))%>%
  select(-violation)%>%
  group_by(date, subject_race, subject_sex, red_stop, seat_belt, fin_resp, license, speeding)%>%
  summarise(n_individuals = n())
```

```{r Barplot Function, echo=FALSE, results='hide'}
barplot <- function(red_stop_m, red_stop_f, seat_belt_m, seat_belt_f, fin_resp_m, fin_resp_f, license_m, license_f, speeding_m, speeding_f){
  y <- c(0.5, 0.5)
  x <- c(0, 1.1)
  #grid.newpage()
  citations <- c("red_stop", "seat_belt", "fin_resp", "license", "speeding")
  male <- c(red_stop_m, seat_belt_m, fin_resp_m, license_m, speeding_m)
  max_male <- max(male)
  female <- c(red_stop_f, seat_belt_f, fin_resp_f, license_f, speeding_f)
  max_female <- max(female)
  colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
  for (i in 1:5){
    male_scale <-  male[i]/(2*max(male))
    female_scale <- female[i]/(2*max(female))
    grid.rect(x=0.2*i-0.1, y = 0.5+0.5*male_scale, height = male_scale, width = 0.199, gp = gpar(fill = colors[i]))
    grid.rect(x=0.2*i-0.1, y = 0.5-0.5*female_scale, height = female_scale, width = 0.199, gp = gpar(fill = colors[i]))
    #grid.text(citations[i], x = 0.2*i-0.1, y = 0.52)
  }
   grid.lines(x, y, gp = gpar(lwd = 1, col = "black"))
   #grid.text ("M", x = 0.9, y = 0.9)
   #grid.text ("F", x = 0.9, y = 0.1)
}
```

```{r slay, echo=FALSE, results='hide'}
bar_run <- function(plot_race,plot_year){
  new_df <- citations_htx_bar[which(citations_htx_bar$date == plot_year &
                                      citations_htx_bar$subject_race==plot_race), ]
  male_df <- new_df[which(new_df$subject_sex == "male"),]
  female_df <- new_df[which(new_df$subject_sex == "female"),]
  red_stop_m <- sum(male_df$red_stop == 1)
  red_stop_f <- sum(female_df$red_stop == 1)
  seat_belt_m <- sum(male_df$seat_belt == 1)
  seat_belt_f <- sum(female_df$seat_belt == 1)
  fin_resp_m <- sum(male_df$fin_resp == 1)
  fin_resp_f <- sum(female_df$fin_resp == 1)
  license_m <- sum(male_df$license == 1)
  license_f <- sum(female_df$license == 1)
  speeding_m <- sum(male_df$speeding == 1)
  speeding_f <- sum(female_df$speeding == 1)
  barplot(red_stop_m, red_stop_f, seat_belt_m, seat_belt_f, fin_resp_m, fin_resp_f, license_m, license_f, speeding_m, speeding_f)
}
#bar_run("white", 2014)  
```

```{r Circles, echo=FALSE, results='hide'}

#DRAW CIRCLE FUNCTION DONE

#given frequency, plot a properly scaled circle in the center of the viewpoint
#max_cite is the citation count of the race, year with the most citations
#found by max(cite_counts_list)
draw_circle <- function(count, max_count = max_cite){
  my_circle <- circleGrob(name = "my_circle",
                          x = 0.5, y = 0.5, r = .8*count/(max_count), #radius scale needs to be fixed and them im done
                          gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.draw(my_circle)
}

# draw_circle(6, 10)
# popViewport()

```

```{r, echo=FALSE, results='hide'}
legend <- function(citations, colors){
  grid.rect(x = 0, y= 0, just = c("left", "bottom"), width = 1, height = .9)
  grid.text("LEGEND",
          x = .5, y = 0.915, just = c("center", "bottom"),
          gp = gpar(fontsize = 10, fontface = "bold"))

  #citations + corresponding colors
  #citations <- c("Red/Stop", "Seat Belt", "Financial", "License", "Speeding")
  #colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
  ys <- c(0.74, 0.65, 0.56, 0.47, 0.38)
  grid.text("Citation Type", x= 0.05, y= 0.87, just = c("left", "top"), gp = gpar(fontsize=8, fontface="bold"))
  for (i in 1:5){
    grid.rect(x = 0.17, y = ys[i], just = c("left", "bottom"), height = 0.077, width = 0.15, 
              gp = gpar(col = "black", fill=colors[i]))
    grid.text(citations[i], x = 0.37, y = ys[i]+0.038, just = c("left", "center"), gp = gpar(fontsize=7))
  }

  grid.text("Citation Frequency", x = 0.04, y = 0.34, just = c("left", "top"), gp = gpar(fontsize=8, fontface="bold"))
  grid.text("Compared to other (race, year)s", x = 0.037, y = 0.3, just = c("left", "top"), gp = gpar(fontsize=5))
  #comparatively less citations for race that year 
  #comparatively more citations

  grid.circle(x = 0.15, y = 0.233, r = 0.04, gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.text("Less Citations", x = 0.3, y = 0.233, just = c("left", "center"), gp = gpar(fontsize=7))
  grid.text("Overall", x = 0.43, y = 0.2, just = c("left", "center"), gp = gpar(fontsize=7))

  grid.circle(x = 0.18, y = 0.11, r = 0.175, gp = gpar(col = "black", fill = "#BEBEBE66", lty = 1, lwt = 2))
  grid.text("More Citations", x = 0.4, y = 0.11, just = c("left", "center"), gp = gpar(fontsize=7))
  grid.text("Overall", x = 0.53, y = 0.075, just = c("left", "center"), gp = gpar(fontsize=7))
}
```

```{r Putting Together the Killer Plot, echo=FALSE, message=FALSE}
# PUT IT ALL TOGETHER
grid.newpage()

#legend viewport 
pushViewport(viewport(
  x = 0.84, y = 0.33, just = c("left", "bottom"),
  width = 0.16, height = 0.57
))

#plot legend
citations <- c("Red/Stop", "Seat Belt", "Financial", "License", "Speeding")
colors <- c("lightcyan","lightpink","moccasin","skyblue1","thistle1")
legend(citations, colors)

popViewport()

#main viewport
pushViewport(viewport(
  x = 0.1, y = 0.1, just = c("left", "bottom"),
  width = 0.8, height = 0.9
))
#axis
grid.lines(c(1.75/8, 0.9), c(1.5/8, 1.5/8))
grid.lines(c(1.75/8, 1.75/8), c(1.5/8, 0.9))
#grid.lines(c(0.1, 0.9), c(0.5, 0.5))
grid.text("Killer Plot",
          x = 1.75/8, y = 0.925, just = c("left", "bottom"),
          gp = gpar(fontsize = 18)
)
pushViewport(viewport(
  x = 0.1, y = 0.1, just = c("left", "bottom"),
  width = 0.8, height = 0.8
))
#grid.rect(x=0.5, y=0.5, width = 1, height=1)
rows <- 8
cols <- 6
ly <- grid.layout(rows, cols)
pushViewport(viewport(layout = ly))
#makes a viewpoint for each

# vectors to iterate through for different parameters
races <- unique(citations_htx_bar$subject_race) # races
years <- unique(citations_htx_bar$date) # dates

# set row labels
for (i in 1:(rows-1)) {
  pushViewport(viewport(layout.pos.row=i, layout.pos.col = 1))
  grid.text(years[i])
  popViewport()}

# set column labels
for (j in 2:cols){
  pushViewport(viewport(layout.pos.row=8, layout.pos.col = j))
  if (races[j-1]=="asian/pacific islander"){
    grid.text("AAPI", gp = gpar(fontsize = 12))
  }
  else{
    grid.text(races[j-1], gp = gpar(fontsize = 12))
  }
  popViewport()
}

for (i in 1:(rows-1)) {
  for (j in 2:cols) {

    # push main grid Viewport
    pushViewport(viewport(layout.pos.row=i, layout.pos.col = j))

    # push small viewport
    pushViewport(viewport(width=0.75, height=0.75, just="centre"))

    # make bar plot
    bar_run(races[j-1], years[i])

    if (j==2){
      if (i==1){
        grid.text("M", x = 0.1, y = 0.9, gp=gpar(fontsize=8))
        grid.text("F", x = 0.1, y = 0.1, gp=gpar(fontsize=8))
      }
    }

    # make circle
    count <- pull(citations_htx_circle[citations_htx_circle['subject_race']==races[j-1] &                          citations_htx_circle['date']==years[i],"n_individuals"])[1]
    #max_count <- sum(citations_htx_circle$n_individuals)
    max_count <- max(citations_htx_circle$n_individuals)
    draw_circle(count, max_count)

    # pop small viewport
    popViewport()

    # pop main grid viewport
    popViewport()
  }
}
```
To truly analyze the key factors of our dataset–how citations change with regards to time, demographics, and location, we created a killer plot that could visually represent a combination of all these aspects of our data. We created a way to visualize citation breakdown by sex, race, year, and district all in one plot. NOTE: this plot is the overall killer plot, and is not subdivided by district. We created one in Shiny which cannot be inserted into a PDF in this manner. This plot here encompasses all districts. 
On the x-axis, we have the different races in our dataset, while the y-axis are different years. For each (race,year) pair, we created a bar plot where bars above the line represent citations for men, while below the midline represent citations for women. We harvested the free text field of “Citation Type” from our dataset to split the citations into 5 main categories–running a red light or stop sign, failure to wear a seatbelt, failure to establish financial responsibility, operating a vehicle with an invalid driver's license, and driving over the speed limit. The height of the bars represent the frequency of that citation (compared to the other citation types) for that race that year.  The gray circles on the killer plot have a radius that is relative to the greatest number of citations for any race, year combination. This plot is comparative by nature and does not have any objective representations of exact numbers of citations, just how they compare between different years and demographics. All of our important categorical information is represented here in one single killer plot.

# Citation-Specific Analysis
Following exploration, our team had a more thorough understanding of the dataset and potential areas that we could investigate for bias. Specifically, our team took interest in the role of race in citations and sought to further analyze trends related to subject race.

First, looking at our data, our team noticed that there were 5 main types of violations: speeding, invalid license, failure to establish financial responsibility, failure to wear a seat belt, and running a stop sign/red light. We wanted to analyze the racial breakdown of each type of citation in order to see if any racial group disproportionately received any type of citation.

```{r Table of race by citation, echo=FALSE}
speeding_race <- mini_htx %>% filter(grepl("SPEEDING", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals)) %>%
  rename("Subject Race"="subject_race", "Speeding"="n_individuals")

# Invalid License Citations
license_race <- mini_htx %>% filter(grepl("LICENSE", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))%>%
  rename("Subject Race"="subject_race", "Invalid License"="n_individuals")

# Failure to establish Financial Responsibility Citations
financial_race <- mini_htx %>% filter(grepl("FINANCIAL RESPONSIBILITY", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))%>%
  rename("Subject Race"="subject_race", "Financial Responsibility"="n_individuals")

# seat belt citations
seat_belt_race <- mini_htx %>% filter(grepl("SEAT BELT", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))%>%
  rename("Subject Race"="subject_race", "Seat Belt"="n_individuals")

# Running a stop light/red light
stop_light_race <- mini_htx %>% 
  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))%>%
  rename("Subject Race"="subject_race", "Stop Sign/Red Light"="n_individuals")

citation_race <- inner_join(speeding_race, license_race, by = "Subject Race")
citation_race <- inner_join(citation_race, financial_race, by = "Subject Race")
citation_race <- inner_join(citation_race, seat_belt_race, by = "Subject Race")
citation_race <- inner_join(citation_race, stop_light_race, by = "Subject Race")
kable(citation_race)

```
For nearly all types of citations, when considering subjects with defined races, white individuals received the most citations, followed by black and AAPI individuals; this aligns with the racial breakdown of the city of Houston.      
However, black individuals received the most citations for failure to establish financial responsibility, which refers to the inability of the subject to provide proof of insurance. Typically, this citation should only be issued given that the subject has committed some other infraction that necessitates police interaction and request for proof of insurance. So, suspicion may be raised around citations that record failure to establish financial responsibility as the sole infraction, as the officer has not recorded any indication of why the individual was pulled over in the first place. Citations that only reference failure to establish financial responsibility may be useful in identifying possible racial bias, as the officer may have pulled over the individual based on their appearance/race, since they did not indicate any other offense on the citation. This led our team to investigate what proportion of citations for individuals of all races resulted solely from failure to establish financial responsibility.

```{r Financial Responsibility by Race Table, echo=FALSE}
# black
black <- mini_htx[mini_htx$subject_race=='black',]
black_total <- nrow(black)

blk_fin <- black %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
blk_perc <- blk_fin[1]/black_total

# white
white <- mini_htx[mini_htx$subject_race=='white',]
white_total <- nrow(white)

white_fin <- white %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
white_perc <- white_fin[1]/white_total

# aapi
aapi <- mini_htx[mini_htx$subject_race=='asian/pacific islander',]
aapi_total <- nrow(aapi)
aapi_fin <- aapi %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
aapi_perc <- aapi_fin[1]/aapi_total

# unknown
unknown <- mini_htx[mini_htx$subject_race=='unknown',]
unknown_total <- nrow(unknown)

unknown_fin <- unknown %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
unknown_perc <- unknown_fin[1]/unknown_total

# NA
nas <- mini_htx[is.na(mini_htx$subject_race),]
nas_total <- nrow(unknown)

nas_fin <- nas %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
nas_perc <- nas_fin[1]/nas_total

# Other
other <- mini_htx[mini_htx$subject_race=="other",]
other_total <- nrow(unknown)

other_fin <- other %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
  summarise(n_individuals = n()) %>%
  arrange(desc(n_individuals))
other_perc <- other_fin[1]/other_total

all_perc <- data.frame(c(blk_perc, white_perc, aapi_perc, unknown_perc, other_perc, nas_perc))
colnames(all_perc) <- c("Black", "White", "AAPI", "Unknown", "Other", "Unrecorded (N/A)")
kable(all_perc)

```
Of all recorded races, the proportion of total citations that mention failure to establish financial responsibility as the sole violation is the highest for individuals of unrecorded race (N/A), followed by black and white individuals and unknown, asian/pacific islanders, and other. It is worth noting that the proportion of financial responsibility citations for unknown race, black, and white individuals were about eighteen, ten, eight times the proportion of these citations for those of AAPI race, respectively. Furthermore, the extremely low proportion of individuals of "Other" race receiving citations solely for failure to establish financial responsibility can be attributed to the extremely low number of citations with race marked as "Other."
After receiving this data, our team ran hypothesis tests at a 95% confidence interval in order to identify which races have a proportion of citations issued solely for failure to establish financial responsibility that is higher than the same proportions for other races. The proportion of these citations for unrecorded race (N/A) was greater than the proportion for all other races by a statistically significant amount. Additionally, the proportion for black individuals was greater than those of white, unknown, and AAPI race by a statistically significant amount. White individuals also had a proportion that was greater than that of unknown and AAPI race.  
Excluding unrecorded race, these findings raise suspicion for bias against black individuals as compared to all other races and for white individuals against individuals of AAPI and unknown race. A high proportion of individuals with unrecorded race receiving citations solely for failure to establish financial responsibility can also be suspicious, as this indicates that officers failed to, whether consciously or unconsciously, record the race of these individuals.

Next, we wanted to investigate if different races and sexes received speeding citations equally, or if certain groups were given citations with different frequencies or for different severity of speeding. We started by averaging the entire dataset to create a baseline.

```{r Speeding Citations, echo=FALSE, message=FALSE}
all_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
  filter(posted_speed >= 0) %>% 
  summarise(mean = mean(diff, na.rm = TRUE), 
            median = median(diff, na.rm = TRUE), 
            sd = sd(diff, na.rm = TRUE))%>%
  rename("Mean"="mean", "Median"="median", "SD"="sd")
race_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
  filter(posted_speed >= 0) %>% group_by(subject_race) %>%
  summarise(total = n(),  ## Function to use internally of summarise only
            mean = mean(diff, na.rm = TRUE),
            median = median(diff, na.rm = TRUE),
            sd = sd(diff, na.rm = TRUE))%>%
  rename("Subject Race"="subject_race", "Total"="total", "Mean"="mean", "Median"="median", "SD"="sd") 
gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
  filter(posted_speed >= 0) %>% group_by(subject_sex) %>%
  summarise(total = n(),  
            mean = mean(diff, na.rm = TRUE),
            median = median(diff, na.rm = TRUE),
            sd = sd(diff, na.rm = TRUE)) %>%
  rename("Subject Sex"="subject_sex")
race_gender_speed <- mutate(mini_htx, diff = speed - posted_speed) %>% 
  filter(posted_speed >= 0) %>% group_by(subject_race, subject_sex) %>%
  summarise(total = n(),
            mean = mean(diff, na.rm = TRUE),
            median = median(diff, na.rm = TRUE),
            sd = sd(diff, na.rm = TRUE)) 
race_gender_speed <- race_gender_speed %>%
  filter_at("subject_sex", all_vars(!is.na(.))) %>%
  rename("Subject Sex"="subject_sex", "Subject Race" = "subject_race", "Total"="total", "Mean"="mean", "Median"="median", "SD"="sd")
kable(all_speed)
kable(race_speed)
options(warn=-1)
```
For the breakdown by race, there is not much evidence of bias on this level. 
One thing to note is the ‘unknown’ category. Police officers fill out these citation reports without asking the subject about their race(s) or ethnicity(s). All of these reported races are from the perspective of the police officers and therefore it is difficult to know the true breakdown. ‘Unknown’ is different from N/A or unrecorded, because the officers did fill out the race section, but itr is an admission that they don’t actually know the race of the subject. 
Here, the unknown could be evidence of bias. We see that for people the officers are unsure about (and those they didn’t ask to clarify), they are given citations for slower speeds (instead of being let off with a warning, for example). This is unproved and could also be explained by the smaller number of unknowns in the sample. 
```{r by gender, echo=FALSE}
kable(gender_speed)
```
Here we see that men and women are ticketed for roughly the same speeds, but there are far more men ticketed than women in our dataset. This is not equal to the proportion of men and women in Houston, so this is either evidence that men speed more often than women (due to higher risk tolerances or different driving behaviors), or that officers are more likely to give men citations and let women off with a warning, or that women are simply caught speeding less often. 
```{r by race and gender, echo=FALSE}
kable(race_gender_speed)
```
When analyzing both race and gender, the key observation to note is that black women are given speeding citations at a very similar rate as black men, while women of all other races have speeding citation counts of nearly half of their male counterparts. Assuming that women of all races speed roughly the same amount, this break in the pattern could be evidence of bias, as it deviates so much from the norm.

# Conclusion
In the end, we found preliminary evidence of bias that can be used as a starting point for future investigation and can potentially guide future policies and interventions. These preliminary biases were seen in the differences of proportions by race for failure to show financial responsibility as the only citation, particularly for black individuals. Additionally, black women were given speeding citations at roughly the same rate as black men, while the rate of speeding citations for women of all other races was consistently around half of their male counterparts. These peculiarities warrant further investigation, as our analysis unfortunately has some key limitations. Most importantly, our dataset has a lot of messy and missing data. Many fields were free text which is very prone to human error. Many observations have missing data in the race field, and we were not able to discern if there was any pattern to the missingness or if it was missing at random. This is important because if a police officer were to intentionally leave the race category blank when giving a questionable citation, their bias could be obfuscated from our analysis. The only way to definitively diagnose bias is with more data with less missing values in the race category. If that dataset did exist, then the severity of bias could change. It could either become stronger, which would corroborate our findings, or disappear completely, which would mean that what we found was simply an anomaly of the pattern of missingness in this dataset. 
Despite this limitation, we believe dedicating time and effort to projects like this is important on chance that any bias is illuminated. Holding our police systems accountable and ensuring accountability is fundamentally important to ensure our societal infrastructure functions properly and fairly for all people. It is an essential step to ensure equal treatment and protection under the law for all individuals, regardless of their demographic.

### 2023-04-20 Final Presentation.Rmd
hopefully last push!

@@
 ---
-title: "Final Presentation"
+title: "Investigating Bias in Houston Police Stop Citations"
 author: "Siddhi Narayan, Grace Wang, Virginia Baskin"
-date: "2023-04-17"
+date: "2023-04-18"
 output: 
@@
 
-## Citations Per Square Mile | Statistics on citations per square mile:
-```{r csqm_summry, echo = FALSE, message = FALSE}
-knitr::kable(cqm_summary)
-```
-
 # Killer Plot!
@@
 ## Killer Plot
-<iframe src="http://127.0.0.1:5501" width="100%" height="500px"></iframe>
+<iframe src="http://127.0.0.1:4965" width="100%" height="500px"></iframe>
 
@@
 ## Citation Type Racial Breakdown {.smaller}
-Speeding Citations by Race:
-```{r speeding_race, echo = FALSE, message = FALSE}
-knitr::kable(speeding_race)
-```
-##
-Invalid License Citations by Race:
-```{r license_race, echo = FALSE, message = FALSE}
-knitr::kable(license_race)
-```
 
-##
-Failure to Establish Financial Responsibility Citations by Race:
-```{r financial_race, echo = FALSE, message = FALSE}
-knitr::kable(financial_race)
-```
-
-##
-Seat Belt Citations by Race:
-```{r seat_belt_race, echo = FALSE, message = FALSE}
-
-knitr::kable(seat_belt_race)
-```
-
-##
-Running a Red Light/Stop Sign Citations by Race:
-```{r stop_light_race, message = FALSE}
-knitr::kable(stop_light_race)
-```
-
-## CHANGE THE NAMES
 ```{r mega table, message = FALSE}
@@
 
-## Failure to Establish Financial Responsibility {.smaller}
+## Failure to Establish Financial Responsibility % {.smaller}
 ```{r fin_resp, echo = FALSE, message = FALSE}
@@
 
-## Average Speeding Amounts {.smaller}
-All People (baseline):
-```{r all_speed, echo = FALSE, message = FALSE}
-knitr::kable(all_speed)
-```
-##
-By Race:
+## Average Speeding Amounts | By Race:
+
 ```{r race_speed, echo = FALSE, message = FALSE}
@@
 ```
-##
-By Gender:
+
+## Average Speeding Amounts | By Sex: 
 ```{r gender_speed, echo = FALSE, message = FALSE}
@@
 ##
+By Race and Sex: 
 ```{r race_gender_speed, echo = FALSE, message = FALSE}

### 2023-04-20 FinalReport.Rmd
hopefully last push!

@@
 date: "2023-04-18"
+header-includes:
+    - \usepackage{setspace}\doublespacing
 output: pdf_document
@@
 
-mini_htx <- read_csv("Datasets/mini_htx.csv")
-# full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
+#mini_htx <- read_csv("Datasets/mini_htx.csv")
+full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv") #was full_htx
 # census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
-# mini_htx <- full_htx[seq(1,2000000,3),]
+mini_htx <- full_htx[seq(1,2000000,3),]
 mini_htx <- mini_htx %>% rename("Beats" = "beat")
@@
 ## Problem and Importance
+
 For our project, we wanted to explore vehicular police stop data from Houston, Texas to investigate and identify any potential biases against particular populations.
@@
 ## Primary Dataset: Police Citations in Houston, Texas
+
 The primary dataset that we used was obtained from the Stanford Open Policing Project, which is a project based in Stanford University that works on collecting and standardizing data on vehicle and pedestrian stops from law enforcement departments across the country.
 For this project, we looked at the data based in Houston.
+
 This dataset consists of a collection of 2,045,972 stops, providing us with an extensive amount of data to work with.
@@
 We also have information on the police beat and police district where the stop was performed.
+
 Additionally, if the stop was a vehicular stop, the dataset includes data on the speed of the vehicle and the speed limit of the area, which can be helpful in understanding the extent to which traffic violations are being targeted by law enforcement officers.
@@
 ### Census Data
+
 One of the supplementary datasets that our team used was the Census Data of the City of Houston from 2021, which we accessed through the United States Census Bureau website.
@@
 ### GIS/Beats Data
+
 Another supplementary dataset our team utilized was GIS Data for the City of Houston Police Beats.
@@
 # Exploration
+
 The purpose of our exploration was to understand and contextualize the large primary dataset and investigate any trends that might exhibit potential for bias.
 We did this by examining a few different factors: distribution  of citations over time, by demographic, and by location.
+
 For distribution of citations over time, we aimed to see if there were any discrepancies between citations for each race over time, and generally see how citations differed from 2014-2019.
 We also looked at the frequency of citations per hour to see if there were any particular patterns, as well as the miles per hour over speed limit per year.
+
 For distribution by demographic, we examined the data in several ways: a waffle chart of citation by race, a table comparing citation proportions for each race to their population proportion in Houston, and a plot of MPH over Speed Limit for Race/Gender.
+
 For distribution by location, we examined this through a population plot of citations by sex and district, as well as a table that displays the number of citations per square mile.
+
 Finally, we combined our investigation of time, demographic, and location in our killer plot, which gave us insight into the distribution of citations by race, gender, and district over time.
@@
 ```
+
 First, we evaluated the consistency of the distribution of citations between races over the years 2014-2020.
@@
 This is further supported by the fact that there is a lull in citations during the super late and super early hours when there is typically less travel.
+
 Another possible explanation for the peak around 3 pm is that it is related to school schedules.
@@
 ```
+
 This representation of the percentage of citations by race across all years 2014-2020 provides a visual and geometric way to understand the distribution of citations across different racial groups.
 The graph shows that white individuals have the highest percentage of citations, followed by black individuals, and then Asian/Pacific Islanders and unknown races.
+
 Following this plot, our team determined that we would like to compare the distribution of citations by race to the demographics of Houston to see if any racial groups receive a disproportionate amount of citations as compared to their representation within the city of Houston.
@@
 ```
+
 This table shows that the citation proportion for White people is higher than their population proportion in Houston, indicating that they may be disproportionately cited.
@@
 However, it's essential to keep in mind that the exact reasons for these disparities are not clear, and there may be several factors contributing to them, such as socioeconomic factors or geographical location.
+
 It's also worth noting that the "other" population proportion is significantly higher than the citation proportion.
@@
 ```
+
 Moving forward, we chose to visualize MPH over Speed Limit for Citation by subject race and sex in order to identify any discrepancies between speeding amounts that warrant a ticket for different races and genders.
@@
 ```
+
 To begin our investigation by location, we made a pyramidal graph of citation count by sex for each district.
@@
 While the core reason for the discrepancy is unknown, the unequal distribution is clear to see.
+
 An interesting feature is that Districts 21 and 23 have significantly less citations overall, and that is because they are Hobby airport and IAH airport, respectively.
 Since it is not residential and technically doesn’t have a population, it is only logical that there are less citations for those districts.
+
 District 2 has the most citations overall, which is one of the most densely populated districts, as it contains Montrose, Greater Heights, West University, and other densely populated neighborhoods within the inner loop of Houston.
@@
 Additionally, we noted that both beats 1A10 and 2A40 are very small in terms of square mileage, but contain two and one police stations, respectively; proximity to police stations could therefore explain the extremely high citations/square mile values for these beats.
+
 Upon receiving the list of 5 police beats with the lowest number of citations, we once again cross-referenced a map of Houston’s police beats to understand the characteristics of these beats.
@@
 #grid.lines(c(0.1, 0.9), c(0.5, 0.5))
-grid.text("Killer Plot",
+grid.text("Citation Breakdown by Sex, Race, & Year",
           x = 1.75/8, y = 0.925, just = c("left", "bottom"),
@@
 ```
+
 To truly analyze the key factors of our dataset–how citations change with regards to time, demographics, and location, we created a killer plot that could visually represent a combination of all these aspects of our data.
@@
 This plot here encompasses all districts.
+
 On the x-axis, we have the different races in our dataset, while the y-axis are different years.
@@
 # Citation-Specific Analysis
+
 Following exploration, our team had a more thorough understanding of the dataset and potential areas that we could investigate for bias.
@@
 ```
+
 For nearly all types of citations, when considering subjects with defined races, white individuals received the most citations, followed by black and AAPI individuals; this aligns with the racial breakdown of the city of Houston.
@@
 ```
+
 Of all recorded races, the proportion of total citations that mention failure to establish financial responsibility as the sole violation is the highest for individuals of unrecorded race (N/A), followed by black and white individuals and unknown, asian/pacific islanders, and other.
@@
 Furthermore, the extremely low proportion of individuals of "Other" race receiving citations solely for failure to establish financial responsibility can be attributed to the extremely low number of citations with race marked as "Other."
+
 After receiving this data, our team ran hypothesis tests at a 95% confidence interval in order to identify which races have a proportion of citations issued solely for failure to establish financial responsibility that is higher than the same proportions for other races.
@@
 White individuals also had a proportion that was greater than that of unknown and AAPI race.
+
 Excluding unrecorded race, these findings raise suspicion for bias against black individuals as compared to all other races and for white individuals against individuals of AAPI and unknown race.
@@
 ```
+
 For the breakdown by race, there is not much evidence of bias on this level.
+
 One thing to note is the ‘unknown’ category.
@@
 ‘Unknown’ is different from N/A or unrecorded, because the officers did fill out the race section, but itr is an admission that they don’t actually know the race of the subject.
+
 Here, the unknown could be evidence of bias.
@@
 This is unproved and could also be explained by the smaller number of unknowns in the sample.
+
 ```{r by gender, echo=FALSE}
@@
 ```
+
 Here we see that men and women are ticketed for roughly the same speeds, but there are far more men ticketed than women in our dataset.
 This is not equal to the proportion of men and women in Houston, so this is either evidence that men speed more often than women (due to higher risk tolerances or different driving behaviors), or that officers are more likely to give men citations and let women off with a warning, or that women are simply caught speeding less often.
+
 ```{r by race and gender, echo=FALSE}
@@
 ```
+
 When analyzing both race and gender, the key observation to note is that black women are given speeding citations at a very similar rate as black men, while women of all other races have speeding citation counts of nearly half of their male counterparts.
@@
 # Conclusion
+
 In the end, we found preliminary evidence of bias that can be used as a starting point for future investigation and can potentially guide future policies and interventions.
@@
 It could either become stronger, which would corroborate our findings, or disappear completely, which would mean that what we found was simply an anomaly of the pattern of missingness in this dataset.
+
 Despite this limitation, we believe dedicating time and effort to projects like this is important on chance that any bias is illuminated.

### 2023-04-20 Presentation.Rmd
hopefully last push!

@@
 #waffle plot: waffle()
+
+```{r, echo=FALSE}
+
+# black
+black <- mini_htx[mini_htx$subject_race=='black',]
+blk_total <- nrow(black)
+
+blk_fin <- black %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
+         summarise(n_individuals = n()) %>%
+         arrange(desc(n_individuals))
+blk_perc <- blk_fin[1]/blk_total
+
+# white
+white <- mini_htx[mini_htx$subject_race=='white',]
+white_total <- nrow(white)
+
+white_fin <- white %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
+         summarise(n_individuals = n()) %>%
+         arrange(desc(n_individuals))
+white_perc <- white_fin[1]/white_total
+
+# aapi
+aapi <- mini_htx[mini_htx$subject_race=='asian/pacific islander',]
+aapi_total <- nrow(aapi)
+
+aapi_fin <- aapi %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
+         summarise(n_individuals = n()) %>%
+         arrange(desc(n_individuals))
+aapi_perc <- aapi_fin[1]/aapi_total
+
+# unknown
+unknown <- mini_htx[mini_htx$subject_race=='unknown',]
+unknown_total <- nrow(unknown)
+
+unknown_fin <- unknown %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
+         summarise(n_individuals = n()) %>%
+         arrange(desc(n_individuals))
+unknown_perc <- unknown_fin[1]/unknown_total
+
+# NA
+nas <- mini_htx[is.na(mini_htx$subject_race),]
+nas_total <- nrow(unknown)
+
+nas_fin <- nas %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
+         summarise(n_individuals = n()) %>%
+         arrange(desc(n_individuals))
+nas_perc <- nas_fin[1]/nas_total
+
+all_perc <- data.frame(c(blk_perc, white_perc, aapi_perc, unknown_perc, nas_perc))
+colnames(all_perc) <- c("black", "white", "aapi", "unknown","N/A")
+all_perc
+```
+Of all recorded races, the proportion of total citations that mention failure to establish financial responsibility as the sole violation is the highest for black individuals.
+This proportion is about 0.01 higher for black individuals compared to white individuals, while aapi individuals and those of unknown race share a similar proportion around 0.0025.
+However, the greatest proportion of total citations that mention failure to establish financial responsibility as the sole violation is attributed to "N/A", which indicates that the officer neglected or failed to record the subject's race.
+
+```{r hypothesis test}
+
+hyptest_finresp <- function(race1, race2, sig){
+  pval <- prop.test(x = c(pull(get(paste0(race1, "_fin"))), pull(get(paste0(race2, "_fin"))))
+                    , n = c(get(paste0(race1, "_total")),
+                            get(paste0(race2,"_total"))), alternative='greater')$p.value
+  if (pval<sig){
+    cat("pval is:", pval, "\nThe null hypothesis is rejected. \n")
+  }
+  if (pval>=sig){
+    cat("pval is:", pval, "\nThe null hypothesis is not rejected. \n")
+  }
+}
+
+for (race in c("white", "aapi", "unknown", "nas")){
+  hyptest_finresp("blk", race, 0.05)
+}
+
+for (race in c("aapi", "unknown")){
+  hyptest_finresp("white", race, 0.05)
+}
+
+```
+
+---
+title: "Final Presentation"
+author: "Siddhi Narayan, Grace Wang, Virginia Baskin"
+date: "2023-04-17"
+output: 
+  ioslides_presentation:
+    smaller: true
+    widescreen: true
+    css: style.css
+runtime: shiny
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, fig.width = 9, fig.height = 5, out.width = "100%")
+source("shiny_plot_setup.R", echo=FALSE)
+source("Tables/Racial_Citation.R", echo = FALSE)
+source("Tables/Racial_Breakdown.R", echo = FALSE)
+source("Tables/FinResp.R", echo = FALSE)
+source("Tables/average_speeding_amt.R", echo = FALSE)
+source("Tables/csqm.R", echo = FALSE)
+```
+
+# Introduction
+
+## Primary Dataset: Citations in HTX
+Insert description
+
+## Secondary Datasets
+* Census Data
+* GIS/Beats Data
+
+## Exploration
+- bullet 
+- point
+
+# Distribution Over Time
+
+## Citation Count by Race over the Years {.smaller}
+```{r Race/Year Line, message=FALSE}
+source("subscripts/line_race_year.R", local=TRUE, print.eval=TRUE)
+line_ry()
+```
+
+## Citation Frequency for Time of Day {.smaller}
+```{r Time of Day Freq, message=FALSE}
+source("subscripts/line_timeofcites.R", local=TRUE, print.eval=TRUE)
+time_line()
+```
+
+# Distribution by Demographic
+
+## Citation by Race {.smaller}
+```{r Waffle, message=FALSE}
+source("subscripts/waffle.R", local=TRUE, print.eval=TRUE)
+waffle()
+```
+
+## MPH over Speed Limit for Race & Gender {.smaller}
+```{r Violin, message=FALSE}
+source("subscripts/violin_speed.R", local=TRUE, print.eval=TRUE)
+violin_speed()
+```
+
+# Distribution by Location
+
+## Citation by Sex by District {.smaller}
+```{r Pyramid, message = FALSE}
+source("subscripts/citation_pyramid.R", local = TRUE, print.eval = TRUE)
+pyramid_cites()
+```
+
+## Citations Per Square Mile | Top 5 Beats ranked by citations/square mile:
+```{r csqm_top, echo = FALSE, message = FALSE}
+knitr::kable(csqm_top)
+```
+
+## Citations Per Square Mile | Bottom 5 Beats ranked by citations/square mile
+```{r csqm_bottom, echo = FALSE, message = FALSE}
+knitr::kable(csqm_bottom)
+```
+
+## Citations Per Square Mile | Statistics on citations per square mile:
+```{r csqm_summry, echo = FALSE, message = FALSE}
+knitr::kable(cqm_summary)
+```
+
+## Killer Plot
+<iframe src="http://127.0.0.1:5501" width="100%" height="500px"></iframe>
+
+# Analysis
+
+## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
+```{r racial_citation_breakdown, echo = FALSE, message = FALSE}
+knitr::kable(htx)
+```
+
+## Citation Type Racial Breakdown {.smaller}
+Speeding Citations by Race:
+```{r speeding_race, echo = FALSE, message = FALSE}
+knitr::kable(speeding_race)
+```
+##
+Invalid License Citations by Race:
+```{r license_race, echo = FALSE, message = FALSE}
+knitr::kable(license_race)
+```
+
+##
+Failure to Establish Financial Responsibility Citations by Race:
+```{r financial_race, echo = FALSE, message = FALSE}
+knitr::kable(financial_race)
+```
+
+##
+Seat Belt Citations by Race:
+```{r seat_belt_race, echo = FALSE, message = FALSE}
+
+knitr::kable(seat_belt_race)
+```
+
+##
+Running a Red Light/Stop Sign Citations by Race:
+```{r stop_light_race, message = FALSE}
+knitr::kable(stop_light_race)
+```
+
+## Failure to Establish Financial Responsibility {.smaller}
+```{r fin_resp, echo = FALSE, message = FALSE}
+knitr::kable(all_perc)
+```
+
+## Average Speeding Amounts {.smaller}
+All People (baseline):
+```{r all_speed, echo = FALSE, message = FALSE}
+knitr::kable(all_speed)
+```
+##
+By Race:
+```{r race_speed, echo = FALSE, message = FALSE}
+knitr::kable(race_speed)
+```
+##
+By Gender:
+```{r gender_speed, echo = FALSE, message = FALSE}
+knitr::kable(gender_speed)
+```
+##
+```{r race_gender_speed, echo = FALSE, message = FALSE}
+library(kableExtra)
+kable(race_gender_speed) %>% kable_styling(font_size=8)
+``` 
+# Conclusion
+
+## Conclusions
+conclusions
+
+Speeding Citations by Race:
+```{r speeding_race, echo = FALSE, message = FALSE}
+knitr::kable(speeding_race)
+```
+##
+Invalid License Citations by Race:
+```{r license_race, echo = FALSE, message = FALSE}
+knitr::kable(license_race)
+```
+
+##
+Failure to Establish Financial Responsibility Citations by Race:
+```{r financial_race, echo = FALSE, message = FALSE}
+knitr::kable(financial_race)
+```
+
+##
+Seat Belt Citations by Race:
+```{r seat_belt_race, echo = FALSE, message = FALSE}
+
+knitr::kable(seat_belt_race)
+```
+
+##
+Running a Red Light/Stop Sign Citations by Race:
+```{r stop_light_race, message = FALSE}
+knitr::kable(stop_light_race)
+```

### 2023-04-20 shiny_plot_setup.R

@@ -8,12 +8,12 @@ library(ggthemes)
 options(dplyr.summarise.inform = FALSE)
 
 # import data
-# full_htx <- read_csv("/stat405final/Datasets/tx_houston_2023_01_26.csv")
+full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
 census_csv <- read_csv("Datasets/Census Data Houston.csv")
 #invisible()
-mini_htx <- read_csv("Datasets/mini_htx.csv")
+#mini_htx <- read_csv("Datasets/mini_htx.csv")
 #invisible(mini_htx)
-# mini_htx <- full_htx[seq(1,2000000,3),]
+mini_htx <- full_htx[seq(1,2000000,3),]
 mini_htx <- mini_htx %>% rename("Beats" = "beat")
 #beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
 #htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")

### 2023-04-20 FinalReport.Rmd
tweaks and transitions

@@
 ## Distribution Over Time
+First, we began our exploration by looking at the distribution of citations over time.
+We first visualized citations issued to each race over the years 2014-2020 in order to evaluate the consistency of the distribution of citations over time.
 ```{r Race Line Graph by Year, echo=FALSE}
@@
 
-First, we evaluated the consistency of the distribution of citations between races over the years 2014-2020.
 We found that the racial distribution of citations remained fairly consistent: from 2014-2019, the greatest number of individuals issued citations were white, but in 2020, black individuals narrowly surpassed white individuals as the racial group with the most citations of any racial group.
@@
 
+Next in our investigation of citations throughout time, we decided to visualize the number of citations issued throughout the hours of the day.
 ```{r Line graph of number of citations vs. Hour in Day, echo=FALSE}
@@
 
-Following this plot, our team determined that we would like to compare the distribution of citations by race to the demographics of Houston to see if any racial groups receive a disproportionate amount of citations as compared to their representation within the city of Houston.
+Following this visualization, our team determined that we would like to compare the distribution of citations by race to the demographics of Houston to see if any racial groups receive a disproportionate amount of citations as compared to their representation within the city of Houston.
 
@@
 attr(cqm_summary, "title") <- "Summary of Citations per Square Mile"
-kable(cqm_summary)
-kable(csqm_top)
-kable(csqm_bottom)
 
@@
 We hoped that this investigation would help us identify any beats with an abnormal quantity of citations so that we could narrow down the scope of our investigation for bias.
-As a part of our investigation, we found the median, standard deviation, and interquartile range of the number of citations per square mile for all police beats and the 5 beats with the greatest and least number of citations per square mile.
+As a part of our investigation, we first found the median, standard deviation, and interquartile range of the number of citations per square mile for all police beats.
+
+```{r, echo=FALSE}
+kable(cqm_summary)
+
+```
+
+We then found the top 5 police beats in terms of number of citations:
+```{r, echo=FALSE}
+kable(csqm_top)
+
+```
 Upon receiving the list of 5 police beats with the greatest number of citations, we cross-referenced a map of Houston’s police beats to view the location and characteristics of these beats.
@@
 Additionally, we noted that both beats 1A10 and 2A40 are very small in terms of square mileage, but contain two and one police stations, respectively; proximity to police stations could therefore explain the extremely high citations/square mile values for these beats.
+Following this, we found the bottom 5 police beats in terms of number of citations: 
+```{r, echo=FALSE}
+kable(csqm_bottom)
 
+```
 Upon receiving the list of 5 police beats with the lowest number of citations, we once again cross-referenced a map of Houston’s police beats to understand the characteristics of these beats.

### 2023-04-20 FinalReport.Rmd
slay shawty

@@
 #mini_htx <- read_csv("Datasets/mini_htx.csv")
-full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv") #was full_htx
+full_htx <- read_csv("/Users/virginiabaskin/Downloads/tx_houston_2023_01_26.csv")
 # census_csv <- read_csv("/Users/virginiabaskin/Downloads/Census Data Houston.csv")
@@
 ## Distribution Over Time
+
 First, we began our exploration by looking at the distribution of citations over time.
 We first visualized citations issued to each race over the years 2014-2020 in order to evaluate the consistency of the distribution of citations over time.
+
 ```{r Race Line Graph by Year, echo=FALSE}
@@
 Next in our investigation of citations throughout time, we decided to visualize the number of citations issued throughout the hours of the day.
+
 ```{r Line graph of number of citations vs. Hour in Day, echo=FALSE}
